---
lecture: "a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.t-title}
::: {.t-eyebrow}
{{< meta course >}}
:::

# Continuous Moderation

::: {.t-rule}
:::

::: {.t-meta}
Unit {{< meta unit >}} · Chapter {{< meta chapter >}} · Lecture {{< meta chapter >}}{{< meta lecture >}}
:::

::: {.t-attrib}
{{< meta attribution >}}
:::

![]({{< meta unit-art >}}){.t-art}
:::

```{r setup}
#| echo: false
#| message: false

# Installs the replace_path source hook, which rewrites the data path shown on
# the slide without changing the path the render actually reads. Quarto does not
# pick _common.R up on its own; without this line every `replace_path` option in
# this deck is a silent no-op.
source("../../_common.R")

library(tidyverse)
library(easystats)
library(patchwork)
mytheme <- function(sz = 30) {
  theme_gray(base_size = sz) + 
  theme(panel.grid.minor = element_blank())
}
# Make mytheme the default rather than relying on every plot remembering
# `+ mytheme`. Plots built by other packages -- plot(), check_model(),
# modelbased -- never get a manual `+ mytheme` and were rendering at
# ggplot's 11pt default, which reads as tiny next to the 40px slide text.
theme_set(mytheme())
```


# Overview

::: {.section-art}
{{< lif "../../icons/mountains.json" colors=secondary:#2a76dd class=rc >}}
:::

## Motivation

- Last chapter, we learned about multiple regression

- We used this to predict $y$ from multiple $x$ variables

- We accounted for the shared variance among $x$ variables

::: {.fragment}
- We will now learn [moderation]{.b .blue} (i.e., interaction effects)

- The effect of one $x$ ***depends on*** the value of another $x$

- This adds complexity and "contextualization" to models

:::

## Example dataset

Imagine we are doing a study on exercise and energy use

```{r}
#| echo: true
#| collapse: true
#| message: false
#| replace_path: ["../../data/", ""]

#install.packages("marginaleffects")
library(tidyverse)
exercise <- read_csv("../../data/exercise.csv")
exercise
```


## Exploratory analysis

```{r, echo=FALSE}
p1 <- qplot(expenditure, data = exercise, geom = "histogram", bins = 30) + mytheme(24)
p2 <- qplot(duration, data = exercise, geom = "histogram", bins = 30) + mytheme(24)
p3 <- qplot(intensity, data = exercise,geom = "histogram", bins = 30) + mytheme(24)
p4 <- qplot(sex, data = exercise, geom = "bar") + mytheme(24)
p5 <- qplot(program, data = exercise, geom = "bar") + mytheme(24)

(p1 | (p2 / p3) | (p4 / p5))
```

::: {.footer}
Note that this is simulated (i.e., fake) data.
:::

## Moderation questions {.fs80}

**Continuous-by-Continuous Interaction**

- Does the effect of `duration` on `expenditure` depend on `intensity`?

::: {.fragment}
**Continuous-by-Categorical Interaction**

- Does the effect of `duration` on `expenditure` depend on `sex`?
:::

::: {.fragment}
**Categorical-by-Categorical Interaction**

- Does the effect of `program` on `expenditure` depend on `sex`?
:::

::: {.fragment}
*Note that each question is statistically identical to its reverse (i.e., vice versa).*
:::

# Continuous-by-Continuous

::: {.section-art}
{{< lif "../../icons/question.json" colors=secondary:#2a76dd class=rc >}}
:::

## Multiple regression recap

Let's regress a continuous $y$ on two continuous $x$ variables

$$\hat{y} = b_0 + b_1 x_1 + b_2 x_2$$

- $b_0$ (intercept): the predicted $y$ when $x_1=0$ and $x_2=0$

- $b_1$ (partial): the change in $y$ for a change of +1 in $x_1$

- $b_2$ (partial): the change in $y$ for a change of +1 in $x_2$

- $b_1$ controls for $x_2$ but is the same across all values of $x_2$ and thus does not depend on it (and the same for $b_2$ and $x_1$)

## Adding an interaction {.fs80}

Interactions are formed by multiplying predictor variables

The product becomes another predictor with its own slope

$$\hat{y} = b_0 + b_1 x_1 + b_2 x_2 + b_3 \color{red}{x_1 x_2}$$

- $b_0$ (intercept): the predicted $y$ when $x_1=0$ and $x_2=0$

- $b_1$ (simple): the change in $y$ for a change of +1 in $x_1$ (specifically at $x_2=0$)

- $b_2$ (simple): the change in $y$ for a change of +1 in $x_2$ (specifically at $x_1=0$)

- $b_3$ (interaction): two equivalent definitions
    + the change in $b_1$ (i.e., the slope of $x_1$) for a change of +1 in $x_2$
    + the change in $b_2$ (i.e., the slope of $x_2$) for a change of +1 in $x_1$

## Rearranging the formula

$$\hat{y} = b_0 + b_1 x_1 + b_2 x_2 + b_3 x_1 x_2$$

::: {.fragment}
$$\hat{y} = b_0 + (b_1 + b_3 x_2) x_1 + b_2 x_2$$

:::

::: {.fragment}
Thus, the slope of $x_1$ (its multiplier) *depends on* the value of $x_2$

Also, if there is no interaction ($b_3=0$), then $b_1$ is unchanged

:::

::: {.fragment .pv2}
The same is true for the slope of $x_2$ if we rearrange again

$$\hat{y} = b_0 + b_1 x_1 + (b_2 + b_3 x_1) x_2$$

:::


## Interaction and simple effects

```{r}
#| echo: true
#| collapse: true

fit <- lm(expenditure ~ duration + intensity + duration:intensity, data = exercise)
model_parameters(fit)
```

- The interaction effect (duration $\times$ intensity) is significant

- Thus, the effect of `duration` does depend on `intensity`

::: {.footer}
We can also use the shortcut `duration * intensity`
:::

## Predicting energy expenditure {.fs80}

$$\hat{y} = 7.80 - 9.38 x_1 - 0.08 x_2 + 0.39 x_1 x_2$$

- **Intercept:** At 0 duration and 0 intensity, a participant expends 7.8 units

- **Duration Simple Effect:** -9.38 units per duration unit (at 0 intensity)

- **Intensity Simple Effect:** -0.08 units per intensity unit (at 0 duration)

- **Interaction Effect:** Each slope rises 0.39 per unit of the other

::: {.fragment}
The average participant (2 duration, 30 intensity) expends 10.04 units of energy

$$\hat{y} = 7.80 - 9.38(2) - 0.08(30) + 0.39(2)(30) = 10.04$$
:::

## Spotlight analysis {.fs80}

- In order to plot the predictions, we can map one predictor to the $x$-axis

- But representing the other predictor continuously is challenging in 2D

::: {.fragment}
- Instead of trying to use a 3D plot, we usually "spotlight" a few values, e.g.,
    + One SD below the mean $(\bar{x} - s_x)$
    + The mean value $(\bar{x})$
    + One SD above the mean $(\bar{x} + s_x)$

:::
::: {.fragment}
- We plot the effect of one $x$ on $y$ conditional on these values of the other $x$
    + Thus, these are often called "conditional effects plots"
    + You can think of "conditional on" as "depending on"

:::

## Plotting interaction effects

```{r}
#| echo: true
#| eval: false
#| collapse: true

library(modelbased)
plot(estimate_relation(fit, by = c("duration", "intensity=[sd]")))
```

```{r}
#| echo: false

library(modelbased)
plot(estimate_relation(fit, by = c("duration", "intensity=[sd]"))) + mytheme(24) + labs(y = "expenditure")
```

::: {.footer}
The `=[sd]` shortcut spotlights intensity at Mean $\pm$ SD. The lines aren't parallel, so the interaction is non-zero.
:::

## Plotting interaction effects

```{r}
#| echo: true
#| eval: false
#| collapse: true

plot(estimate_relation(fit, by = c("intensity", "duration=[sd]")))
```

```{r}
#| echo: false

plot(estimate_relation(fit, by = c("intensity", "duration=[sd]"))) + mytheme(24) + labs(y = "expenditure")
```

::: {.footer}
You can tell there is a non-zero interaction because the lines aren't parallel.
:::

## Simple slopes

::: {.fs70}
- What is the slope of `duration` at different values of `intensity`?
    + With a non-zero interaction, the simple slopes will differ
    
:::

```{r}
#| echo: true
#| collapse: true

estimate_slopes(fit, trend = "duration", by = "intensity=c(0,15,30,45,60)")
```

::: {.footer}
We could also use the shortcut `"intensity=[sd]"` to specify Mean $\pm$ SD
:::

## Contrast with No Interaction

```{r}
#| echo: true
#| collapse: true

fit2 <- lm(expenditure ~ duration + intensity, data = exercise)
model_parameters(fit2)
```

::: {.fs70}
- A participant exercising at 0 duration and 0 intensity expends -15.6 units
- Each additional unit of duration will expend 2.35 units (controlling `intensity`)
- Each additional unit of intensity will expend 0.71 units (controlling `duration`)
- The model predicts the same effect of duration at all values of intensity
- The model predicts the same effect of intensity at all values of duration
:::

## Plotting partial effects

```{r}
#| echo: true
#| eval: false
#| collapse: true

plot(estimate_relation(fit2, by = c("duration", "intensity=[sd]")))
```

```{r}
#| echo: false

plot(estimate_relation(fit2, by = c("duration", "intensity=[sd]"))) + mytheme(24) + labs(y = "expenditure")
```

::: {.footer}
You can tell there is no interaction because the lines are parallel.
:::

## Plotting partial effects

```{r}
#| echo: true
#| eval: false
#| collapse: true

plot(estimate_relation(fit2, by = c("intensity", "duration=[sd]")))
```

```{r}
#| echo: false

plot(estimate_relation(fit2, by = c("intensity", "duration=[sd]"))) + mytheme(24) + labs(y = "expenditure")
```

::: {.footer}
You can tell there is no interaction because the lines are parallel.
:::

## Simple slopes

- In a model with no interaction, all simple slopes are equal

```{r}
#| echo: true
#| collapse: true

estimate_slopes(fit2, trend = "duration", by = "intensity=[sd]")
```

# Interaction Patterns

::: {.section-art}
{{< lif "../../icons/question.json" colors=secondary:#2a76dd class=rc >}}
:::

## CxC Patterns

::: {.columns}
::: {.column width="33%" .tc .fs70}
**Potentiation**

As one $x$ increases, the slope of the other increases

```{r, echo=FALSE, fig.width=3, fig.height=4}
set.seed(2021)
n <- 100
d <- tibble(
  x1 = rnorm(n, 10, 5),
  x2 = rnorm(n, 0, 1),
  y = 10 + 1.1 * x1 + 0.8 * x2 + 0.6*x1*x2 + rnorm(n,0,6) 
)
fit0 <- lm(y ~ x1 * x2, d)
plot(estimate_relation(fit0, by = c("x1", "x2=c(-1,0,1)"))) +
  mytheme(20) + 
  theme(legend.position = "top") +
  labs(title = NULL)
```

:::
::: {.column width="33%" .tc .fs70}
**Attenuation**

As one $x$ increases, the slope of the other decreases

```{r, echo=FALSE, fig.width=3, fig.height=4}
set.seed(2021)
n <- 100
d <- tibble(
  x1 = rnorm(n, 10, 5),
  x2 = rnorm(n, 0, 1),
  y = 0 + 0.1*x1 - 5*x2 + .2*x1*x2 + rnorm(n,0,2) 
)
fit0 <- lm(y ~ x1 * x2, d)
plot(estimate_relation(fit0, by = c("x1", "x2=c(-1,0,1)"))) +
  mytheme(20) + 
  theme(legend.position = "top") +
  labs(title = NULL)
```
:::
::: {.column width="33%" .tc .fs70}
**Crossover**

As one $x$ increases, the ranking of the other reverses

```{r, echo=FALSE, fig.width=3, fig.height=4}
set.seed(2021)
n <- 100
d <- tibble(
  x1 = rnorm(n, 10, 5),
  x2 = rnorm(n, 0, 1),
  y = 0 + .1*x1 - .01*x2 - .025*x1*x2 + rnorm(n,0,2) 
)
fit0 <- lm(y ~ x1 * x2, d)
plot(estimate_relation(fit0, by = c("x1", "x2=c(-1,0,1)"))) +
  mytheme(20) + 
  theme(legend.position = "top") +
  labs(title = NULL)
```
:::
:::

# Centering with Interactions

::: {.section-art}
{{< lif "../../icons/question.json" colors=secondary:#2a76dd class=rc >}}
:::

## Centering interactions

- Recall that the simple effects are when the other $x=0$
    + This means that their $p$ values only apply in that region
    
- So, if $x=0$ isn't meaningful, we can use centering
    + Now they are the slopes when the other $x=\bar{x}$
    
```{r}
#| echo: true
#| collapse: true

exercise$duration_c <- center(exercise$duration)
exercise$intensity_c <- center(exercise$intensity)
```

## Centering interactions

```{r}
#| echo: true
#| collapse: true

fit3 <- lm(expenditure ~ duration_c * intensity_c, data = exercise)
model_parameters(fit3)
```

::: {.fs70}
- A participant with average duration and intensity expends 10.00 units

- Each added unit of duration expends 2.29 units of energy (at average intensity)

- Each added unit of intensity expends 0.71 units of energy (at average duration)

- The slope of duration will increase by 0.39 for each unit of intensity, and the slope of intensity will increase by 0.39 for each unit of duration
:::

## Plotting centered interactions

```{r}
#| echo: true
#| eval: false
#| collapse: true

plot(estimate_relation(fit3, by = c("duration_c", "intensity_c=[sd]")))
```

```{r}
#| echo: false

plot(estimate_relation(fit3, by = c("duration_c", "intensity_c=[sd]"))) + mytheme(24) + labs(y = "expenditure")
```


## Plotting centered interactions

```{r}
#| echo: true
#| eval: false
#| collapse: true

plot(estimate_relation(fit3, by = c("intensity_c", "duration_c=[sd]")))
```

```{r}
#| echo: false

plot(estimate_relation(fit3, by = c("intensity_c", "duration_c=[sd]"))) + mytheme(24) + labs(y = "expenditure")
```

## Standardizing coefficients

- Don't forget that standardizing includes centering!

```{r}
#| echo: true
#| collapse: true

model_parameters(fit, standardize = "refit")
```

::: {.fs70}
- Increasing duration by 1 SD expends another 0.08 SD (at *average* intensity)

- Increasing intensity by 1 SD expends another 0.26 SD (at *average* duration)

- Increasing duration by 1 SD increases the effect of intensity by 0.07 SDs and<br> increasing intensity by 1 SD increases the effect of duration by 0.07 SDs

:::

## Recommendations

- Center your interacting continuous predictors

- Use domain knowledge to guide selection of interactions

- Plot your interaction effects to aid in communication

- Don't include an interaction without its simple effects

- Don't confuse simple effects and partial effects
