---
lecture: "a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.t-title}
::: {.t-eyebrow}
{{< meta course >}}
:::

# Simple Regression

::: {.t-rule}
:::

::: {.t-meta}
Unit {{< meta unit >}} · Chapter {{< meta chapter >}} · Lecture {{< meta chapter >}}{{< meta lecture >}}
:::

::: {.t-attrib}
{{< meta attribution >}}
:::

![]({{< meta unit-art >}}){.t-art}
:::

```{r setup}
#| echo: false
#| message: false

# Installs the replace_path source hook, which rewrites the data path shown on
# the slide without changing the path the render actually reads. Quarto does not
# pick _common.R up on its own; without this line every `replace_path` option in
# this deck is a silent no-op.
source("../../_common.R")

library(tidyverse)
library(easystats)
library(patchwork)
mytheme <- function(sz = 30) {
  theme_gray(base_size = sz) + 
  theme(panel.grid.minor = element_blank())
}
# Make mytheme the default rather than relying on every plot remembering
# `+ mytheme`. Plots built by other packages -- plot(), check_model(),
# modelbased -- never get a manual `+ mytheme` and were rendering at
# ggplot's 11pt default, which reads as tiny next to the 40px slide text.
theme_set(mytheme())
```


# Introduction

::: {.section-art}
{{< lif "../../icons/map.json" colors=secondary:#2a76dd class=rc >}}
:::

## Regression as Fancy Correlation

- A correlation quantifies how well a **straight line** represents the relationship between two continuous variables

```{r, fig.width=12, fig.height=5}
#| echo: false
#| replace_path: ["../../data/", ""]

parenthood <- read_csv("../../data/parenthood.csv", show_col_types = FALSE)
p <- ggplot(parenthood, aes(x = dan.sleep, y = dan.grump)) +
  geom_point(size = 3) +
  labs(x = "sleep", y = "grumpiness") + 
  mytheme()
p
```

## Regression as Fancy Correlation

- A correlation quantifies how well a [straight line]{.b .salmon} represents the relationship between two continuous variables

```{r, fig.width=12, fig.height=5}
#| echo: false

p + geom_smooth(
  formula = y ~ x, 
  method = "lm", 
  se = FALSE, 
  color = "salmon",
  linewidth = 2
)
```

## Regression as Fancy Correlation

- But how did we know *where* to draw the line?

- Many different lines could have been drawn...

```{r, fig.width=12, fig.height=5}
#| echo: false

p + 
  geom_abline(intercept = 125, slope = -8.937, linewidth = 2, color = "#7570b3") +
  geom_abline(intercept = 150, slope = -12, linewidth = 2, color = "#1b9e77") +
  geom_abline(intercept = 100, slope = -5, linewidth = 2, color = "#1f78b4")
```

## Regression as Fancy Correlation

- Regression is a way to *select* and *quantify* the "best" line

```{r, fig.width=12, fig.height=5}
#| echo: false

p + geom_smooth(
  formula = y ~ x, 
  method = "lm", 
  se = FALSE, 
  color = "salmon",
  linewidth = 2
)
```

## Defining a Line in General

::: {.columns}
::: {.column}

$$y = \color{blue}{m}x + \color{blue}{b}$$

$y$ and $x$ are variables

$\color{blue}m$ and $\color{blue}b$ are coefficients

- $\color{blue}m$ is the **slope**<br>
    Change in $y$ when $x+1$

- $\color{blue}b$ is the **intercept**<br>
    Value of $y$ when $x=0$ 

:::
::: {.column}
```{r, fig.width=5, fig.height=5}
m <- 4
b <- 2
ggplot() + 
  geom_polygon(aes(x = c(0, 1, 1), y = c(b, b, b + m)), fill = "blue", alpha = 0.5) + 
  geom_vline(xintercept = 0, color = "grey", linewidth = 1.25) +
  geom_abline(intercept = b, slope = m, linewidth = 2, color = "salmon") +
  scale_y_continuous(limits = c(0, 8), breaks = seq(0, 8, 2)) + 
  scale_x_continuous(limits = c(-2, 2)) +
  geom_segment(aes(x = -0.9, y = b, xend = -0.1, yend = b), linewidth = 1, 
               arrow = arrow(length = unit(0.5, "cm")),
               lineend = "square", linejoin = "mitre") +
  annotate("text", x = -1.5, y = b, label = "b", size = 12) +
  annotate("text", x = 1.75, y = 4.15, label = "m", size = 12) +
  ggbrace::stat_brace(
    aes(x = c(1.1, 1.3), y = c(b, b + m)),
    rotate = 90, 
    linewidth = 1, 
    outside = FALSE
  ) +
  labs(x = "x", y = "y") +
  mytheme()
```

:::
:::

## Defining a Regression Line

- If $y$ is the outcome and $x$ is the predictor...

- We can define the regression line in the same way

$$\color{red}{\hat{y}_i} = \color{blue}{b_0} + \color{blue}{b_1}x_i$$

$\color{red}{\hat{y}_i}$ is the model-predicted value of $y$ for observation $i$

$\color{blue}{b_0}$ is the intercept of the line (i.e., $b$ on the previous slide)

$\color{blue}{b_1}$ is the slope of the line (i.e., $m$ on the previous slide)

## Defining Regression Residuals

::: {.columns}
::: {.column}
- Our regression line is defined by the formula:

$$\color{red}{\hat{y}_i} = \color{blue}{b_0} + \color{blue}{b_1}x_i$$

- But many points aren't exactly on the line, so our predictions aren't perfect

:::
::: {.column}
```{r, fig.width=5, fig.height=5}
p + geom_smooth(method = "lm", se = FALSE, color = "salmon", linewidth = 2)
```

:::
:::

## Defining Regression Residuals

- Thus, predictions $(\color{red}{\hat{y}_i})$ differ from the observed values $(y_i)$

- We refer to these differences as the model's [residuals]{.b .pink}

$$
\begin{align}
\color{red}{\hat{y}_i} &= \color{blue}{b_0} + \color{blue}{b_1}x_i\\
y_i &= \color{blue}{b_0} + \color{blue}{b_1}x_i + \color{magenta}{e_i} \\
\color{magenta}{e_i} &= y_i - \color{red}{\hat{y}_i}
\end{align}
$$

## Visualizing Regression Residuals

```{r, fig.width=12, fig.height=6}
fit <- lm(dan.grump ~ dan.sleep, data = parenthood)
p + 
  geom_smooth(method = "lm", se = FALSE, color = "salmon", linewidth = 2) +
  geom_segment(aes(x = dan.sleep, xend = dan.sleep, 
                   y = dan.grump, yend = dan.grump - residuals(fit)),
               linewidth = 0.75, color = "grey30") 
```


## Minimizing Regression Residuals

- If we want the best possible estimates of $y$, we should...
- ...look for the line that minimizes the prediction errors
- Residuals are the prediction errors
- We thus seek to minimize the residuals

::: {.fragment}
- But residuals can be positive or negative...
- So minimizing the sum of the residuals won't work
- We thus minimize the sum of the *squared* residuals
:::

## OLS Regression

- Minimizing the sum of the squared residuals is called OLS
    - This stands for "Ordinary Least Squares" Regression

- So what are the OLS formulas for the slope and intercept?


$$
b_1 = r_{xy}\frac{s_y}{s_x}
\qquad\quad
b_0 = \bar{y} - b_1\bar{x}
$$

- Regression and correlation are thus integrally connected

## Example Dataset

The `water` dataset has information about mortality and water hardness in 61 UK towns

```{r}
#| echo: true
#| message: false
#| replace_path: ["../../data/", ""]

water <- read_csv("../../data/water.csv")
water
```


## Data Exploration

```{r, fig.width=12, fig.height=6}
p1 <- ggplot(water, aes(x = mortality)) + geom_histogram(bins = 10) +
  mytheme()
p2 <- ggplot(water, aes(x = hardness)) + geom_histogram(bins = 10) +
  mytheme()
p3 <- ggplot(water, aes(x = hardness, y = mortality)) + geom_point(size = 2) +
  mytheme()
(p1 / p2) | p3
```

## OLS Regression by Hand

```{r}
#| echo: true
#| collapse: true

x <- water$hardness
y <- water$mortality

b1 <- cor(x, y) * sd(y) / sd(x)
b1

b0 <- mean(y) - b1 * mean(x)
b0
```

$$
\hat{y}_i = 1676.356 - 3.226 x_i
$$


::: {.fragment .fs70}
"We estimate that a town with a water hardness of zero would have a mortality of 1676.356. We estimate that towns with one more unit of water hardness have mortality that is *lower* by 3.226, on average."
:::

## Visualizing our Result

```{r}
#| echo: true
#| eval: false
qplot(x = hardness, y = mortality, data = water, geom = "point") + 
  geom_abline(slope = b1, intercept = b0, color = "salmon")
```

```{r, fig.width=12, fig.height=5}
ggplot(water, aes(x = hardness, y = mortality)) +
  geom_point(size = 3) + 
  geom_abline(slope = b1, intercept = b0, linewidth = 2, color = "salmon") + 
  mytheme()
```


## Regression using lm()

```{r}
#| echo: true
#| collapse: true

fit <- lm(formula = mortality ~ hardness, data = water)
fit
```

- Just like before, the formula is `outcome ~ predictor`

- R refers to $b_0$ as `(Intercept)` and $b_1$ as `hardness`

- This will be helpful when we add multiple predictors

# Regression Inference

::: {.section-art}
{{< lif "../../icons/insights.json" colors=secondary:#2a76dd class=rc >}}
:::

## Uncertainty in the Population

- OLS finds the best coefficients *for the sample*

- But we usually care more about the **population**

::: {.fragment}
- Thus, we *estimate* the population intercept and slope

- There will be some inferential uncertainty in this process
:::

::: {.fragment}
- As before, we can use NHST and/or estimation approaches

- Let's calculate 95% CIs, test statistics, and $p$-values
:::

## Slope Confidence Interval

$$
b_1 \pm \text{SE}(b_1)\times t_{crit}
$$

$$
\text{SE}(b_1) = \frac{s_y}{s_x}\sqrt{\frac{1-r_{xy}^2}{n-2}}
$$

$$t_{crit}\text{ comes from the } t(n-2, 0, 1)\text{ distribution}$$

## Intercept Confidence Interval

$$
b_0 \pm \text{SE}(b_0)\times t_{crit}
$$

$$
\text{SE}(b_0) = \sqrt{\frac{\sum_i(y_i-\hat{y}_i)^2}{n-2}} \sqrt{\frac{1}{n} + \frac{\bar{x}^2}{(n-1)s_x^2}}
$$

$$t_{crit}\text{ comes from the } t(n-2, 0, 1)\text{ distribution}$$

## Confidence Intervals in R

```{r}
#| echo: true
#| collapse: true

fit

confint(fit)
```

- I'll show you the {easystats} way after we cover NHST

## Plotting Regression Results

```{r}
#| echo: true
#| eval: false

library(modelbased)
est <- estimate_relation(fit, by = "hardness")
plot(est, show_data = TRUE)
```

```{r, fig.width=12, fig.height=5}
#| echo: false
#| eval: true

library(modelbased)
est <- estimate_relation(fit, by = "hardness")
plot(est, show_data = TRUE) + mytheme(24)
```

## Hypothesis Testing

**Null:** The true coefficient (in the population) is zero

$$H_0: b = 0$$

**Alternative:** The true coefficient (in the population) is *not* zero

$$H_1: b \neq 0$$

Note that this applies to both intercepts and slopes

## Testing Individual Coefficients

- We can re-use the logic/procedure of the one-sample $t$-test

- We assume the null sampling distribution for $\hat{b}$ is normal

$$
t_{obs} = \frac{b}{\text{SE}(b)}
$$

$$t_{obs} \sim t(\mathit{df}=n-k-1, \mu=0, \sigma=1)$$

::: {.footer}
Note that $k$ is the number of predictors in the model (i.e., 1 for now).
:::

## Statistical Inference in R

```{r}
#| echo: true
#| collapse: true

library(easystats)
model_parameters(fit)
```

"A town with zero water hardness is expected to have a mortality rate of 1676.36, which is significantly different from zero, 95% CI: [1617.73, 1734.98], *p*<.001. For each one-unit increase in water hardness, the model predicts a significant decrease in mortality of 3.23, 95% CI: [--4.20, --2.26], *p*<.001."

::: {.footer}
Note that the $t$ values equal Coefficient / SE (e.g., --3.23/0.48 ≈ --6.66).
:::

# Centering

::: {.section-art}
{{< lif "../../icons/location_pin.json" colors=secondary:#2a76dd class=rc >}}
:::

## Centering a Predictor

- The intercept $(b_0)$ is the value of $\hat{y}$ when $x=0$

- This isn't so useful if $x$ never equals $0$ in the sample

::: {.fragment}
- In such cases, it can be useful to "center" $x$ before modeling

$$x_i^c = x_i - \bar{x}$$

```{r}
xc <- x - mean(x)
```

- The **slope will stay the same** but the intercept will change

- The intercept will become the value of $\hat{y}$ when $x=\bar{x}$

:::

## Centering a Predictor

::: {.columns}
::: {.column .tc}
```{r, fig.width=5, fig.height=3}
ggplot(water, aes(x = hardness, y = mortality)) +
  geom_vline(xintercept = 0, color = "grey", linewidth = 2) +
  geom_point(size = 2.5) + 
  geom_abline(slope = -3.226, intercept = 1676.356, linewidth = 2, color = "salmon") + 
  scale_x_continuous(breaks = seq(0, 120, 30)) +
  mytheme(24)
```

$$b_0 = 1676.356$$

["A town with *zero* water hardness<br>would have 1676.356 mortality."]{.fs70}
:::
::: {.column .tc}
```{r, fig.width=5, fig.height=3}
ggplot(water, aes(x = hardness - mean(hardness), y = mortality)) +
  geom_vline(xintercept = 0, color = "grey", linewidth = 2) +
  geom_point(size = 2.5) + 
  geom_abline(slope = -3.226, intercept = 1524.148, linewidth = 2, color = "salmon") + 
  scale_x_continuous(breaks = seq(-30, 90, 30)) +
  labs(x = "center(hardness)") +
  mytheme(24)
```

$$b_0 = 1524.148$$

["A town with *average* water hardness<br>would have 1524.148 mortality."]{.fs70}
:::
:::

## Centering in R

```{r}
#| echo: true
#| collapse: true

# Two-step process
water$hardness_c <- water$hardness - mean(water$hardness)

fit2a <- lm(formula = mortality ~ hardness_c, data = water)
model_parameters(fit2a)
```
::: {.fragment .pv2}
```{r}
#| echo: true
#| collapse: true

# Or equivalent one-step process
fit2b <- lm(formula = mortality ~ center(hardness), data = water)
model_parameters(fit2b)
```
:::

# Regression Effect Sizes

::: {.section-art}
{{< lif "../../icons/rulers.json" colors=secondary:#2a76dd class=rc >}}
:::

## Standardized Slopes {.fs90}

- It is common to standardize regression slopes (called $\beta$s)

- We can do this by standardizing both $y$ and $x$ and refitting

$$x_i^z = \frac{x_i - \bar{x}}{s_x} \qquad \quad y_i^z = \frac{y_i - \bar{y}}{s_y}$$

$$\hat{y}_i^z = \beta_0 + \beta_1x_i^z$$

- The standardized slope $(\beta_1)$ is interpreted in SD units
- The standardized intercept $(\beta_0)$ will equal/approach zero

## Standardizing in R

```{r}
#| echo: true
#| collapse: true

# Method 1: Fit normally then refit
fit <- lm(mortality ~ hardness, data = water)
model_parameters(fit, standardize = "refit")

# Method 2: Standardize data then fit
fit2 <- lm(mortality ~ hardness, data = standardize(water))
model_parameters(fit2)
```

::: {.fragment .fs70 .tc}
"For each one SD increase in water hardness, we would expect a 0.65 SD<br>*decrease* in mortality, $\beta=-0.65$, 95% CI: $[-0.85, -0.46]$, $p<.001$."
:::

## R-Squared

- $R^2$ is the variance in $y$ explained by the regression model
    + Ranges from 0.0 (0% explained) to 1.0 (100% explained)
    + It is equivalent to squaring the correlation of $y$ and $\hat{y}$
    + It is also called the "coefficient of determination"
    
$$
R^2 = 1 - \frac{\text{SS}_{res}}{\text{SS}_{tot}} = 1 - \frac{\sum_i(y_i - \hat{y}_i)^2}{\sum_i(y_i - \bar{y})^2}
$$

## Adjusted R-Squared

- $R^2$ will never decrease when adding more predictors

- So we often adjust it to penalize adding more predictors

$$
R_{adj}^2 = 1 - \left(\frac{\text{SS}_{res}}{\text{SS}_{tot}} \times \frac{n - 1}{n - k - 1}\right)
$$

- Negative values of $R_{adj}^2$ are possible

- Its interpretation is less straightforward

## Model Performance in R

```{r}
#| echo: true
#| collapse: true

# By hand (with the residuals function shortcut)
R2a <- 1 - sum(residuals(fit)^2) / sum((y - mean(y))^2)
R2a

# By function (recommended)
R2b <- r2(fit, ci = 0.95)
R2b
```

"The regression model in which mortality was regressed on water hardness explained 42.9% of the variance in mortality, R² = 0.429, 95% CI: [0.230, 0.601], Adjusted R² = 0.419."
