---
lecture: "b"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.t-title}
::: {.t-eyebrow}
{{< meta course >}}
:::

# Categorical Predictors

::: {.t-rule}
:::

::: {.t-meta}
Unit {{< meta unit >}} · Chapter {{< meta chapter >}} · Lecture {{< meta chapter >}}{{< meta lecture >}}
:::

::: {.t-attrib}
{{< meta attribution >}}
:::

![]({{< meta unit-art >}}){.t-art}
:::

```{r setup}
#| echo: false
#| message: false

# Installs the replace_path source hook, which rewrites the data path shown on
# the slide without changing the path the render actually reads. Quarto does not
# pick _common.R up on its own; without this line every `replace_path` option in
# this deck is a silent no-op.
source("../../_common.R")

library(tidyverse)
library(easystats)
library(patchwork)
mytheme <- function(sz = 30) {
  theme_gray(base_size = sz) + 
  theme(panel.grid.minor = element_blank())
}
# Make mytheme the default rather than relying on every plot remembering
# `+ mytheme`. Plots built by other packages -- plot(), check_model(),
# ggeffects -- never get a manual `+ mytheme` and were rendering at
# ggplot's 11pt default, which reads as tiny next to the 40px slide text.
theme_set(mytheme())
```


# Introduction

::: {.section-art}
{{< lif "../../icons/question.json" colors=secondary:#2a76dd class=rc >}}
:::

## Categorical variables

- We learned regression with a single continuous predictor

- Today we learn about regression with a categorical predictor

::: {.fragment}
- We use `factor()` to tell R that a variable is categorical

- Our factor puts each observation into one of $g$ groups
:::

::: {.fragment}
- Groups must be **mutually exclusive** and **exhaustive**
    + No observation is in more than one group/category
    + All observations are in a group/category
:::

## Terminology and examples

- [Two groups]{.b .blue}: "binary" or "dichotomous"
    + e.g., $\text{Condition}\in\{\text{Control}, \text{Treatment}\}$
    + e.g., $\text{Gender}\in\{\text{Man}, \text{Woman}\}$

- [Three or more groups]{.b .blue}: "nominal" or "polytomous"
    + e.g., $\text{Nation}\in\{\text{UK}, \text{Germany}, \text{Spain}, \text{France}\}$
    + e.g., $\text{Sect}\in\{\text{Catholic}, \text{Protestant}, \text{Orthodox}\}$
    + e.g., $\text{Time}\in\{\text{Baseline}, \text{Midpoint}, \text{Endpoint}\}$*

::: {.footer}
*Ordered categories can be modeled as nominal or ordinal.
:::

## Code variables

- To use factors as predictors, we need to *quantify* them
    + We turn them into numerical **code variables**

::: {.fragment}
- There are several methods called [coding systems]{.b .blue}
    + Coding systems may change the code and coefficients
    + But they don't change the predictions or model fit
::: 

::: {.fragment}
- All coding systems turn a factor into $g-1$ code variables
    + So if we have 6 groups, we need 5 code variables
:::

# Dichotomous <br>Dummy Coding

::: {.section-art}
{{< lif "../../icons/question.json" colors=secondary:#2a76dd class=rc >}}
:::

## Coding two categories

- Dummy coding uses a value of $0$ or $1$ for each code variable
    + With two categories, we only need one code variable: $C_1$
    + One category gets $C_1=0$ and the other gets $C_1=1$
    + The category assigned to $0$ is called the "reference group"

::: {.columns .tc .fs70}
::: {.column .tc}
*Female as Reference*
<table style="border-collapse: collapse; margin-left: auto; margin-right: auto;">
<tr class="bb">
<th>Sex</th>
<th>C<sub>1</sub></th>
</tr>
<tr>
<td>Female*</td>
<td>0</td>
</tr>
<tr>
<td>Male</td>
<td>1</td>
</tr>
</table>
:::
::: {.column .tc}
*Male as Reference*
<table style="border-collapse: collapse; margin-left: auto; margin-right: auto;">
<tr class="bb">
<th>Sex</th>
<th>C<sub>1</sub></th>
</tr>
<tr>
<td>Female</td>
<td>1</td>
</tr>
<tr>
<td>Male*</td>
<td>0</td>
</tr>
</table>
:::
:::

## Example dataset

```{r}
#| echo: true
#| collapse: true
#| message: false
#| replace_path: ["../../data/", ""]

penguins <- read_csv("../../data/penguins.csv")
penguins
```

## Dummy coding by hand

- We can use the `recode()` function from {tidyverse}

```{r}
#| echo: true
#| collapse: true

penguins$sex_c1 <- recode(penguins$sex, female = 1, male = 0)
penguins
```

## Regression with a dummy code

- We can regress `body_mass` on the dummy code $C_1$
    
$$\hat{y}_i = b_0 + b_1 C_{1i}$$

- The intercept $b_0$ is the value of $\hat{y}$ when $C_1=0$
    + i.e., the reference group's average outcome value

- The slope $b_1$ is the change in $\hat{y}$ when $C_1+1$
    + i.e., the group difference in average outcome values

## Regression with a dummy code

```{r}
#| echo: true
#| collapse: true

fit <- lm(formula = body_mass ~ sex_c1, data = penguins)
fit
```

- The average body mass of *male* penguins is 4545.68 g

- *Female* penguins weigh 683.41 g less on average

## The factor shortcut (recommended)

- R will automatically generate dummy codes *for factors*
    + It will use the [first level]{.b .blue} as the reference group
    + It will name the slope "factorNonref" (e.g., sexFemale)
    
```{r}
#| echo: true
#| collapse: true

penguins$sex <- factor(penguins$sex, levels = c("male", "female"))
fit2 <- lm(formula = body_mass ~ sex, data = penguins)
fit2
```

## Swapping the reference group

- To use female as the reference group, we just put it first

```{r}
#| echo: true
#| collapse: true

penguins$sex <- factor(penguins$sex, levels = c("female", "male"))
fit3 <- lm(formula = body_mass ~ sex, data = penguins)
fit3
```

- The average body mass of *female* penguins is 3862.27 g

- *Male* penguins weigh 683.41 g more on average

## Parameter significance testing

- Testing is the same as with a continuous predictor

```{r}
#| echo: true
#| collapse: true

model_parameters(fit2)
```

> "The average body mass of female penguins was estimated to be 683.41 grams lower than that of male penguins, 95% CI: [--840.80, --526.02], *t* = --8.54, *p* < .001."

## Relation to Student's t-test {.fs80}

```{r}
#| echo: true
#| collapse: true

fit2b <- t.test(body_mass ~ sex, data = penguins, var.equal = TRUE)
model_parameters(fit2b)
```

- The slope test and Student's *t*-test give the same results
    + The linear model (lm) "subsumes" Student's *t*-test
    + It also assumes equal variances in the two groups

## Effect sizes

- The standardized slope is similar to Cohen's $d$
    
```{r}
#| echo: true
#| collapse: true

model_parameters(fit2, standardize = "refit")
```
::: {.fs70}
> "The average body mass of female penguins was 0.85 standard deviations lower than that of male penguins, 95% CI: [--1.04, --0.65], p<.001."

:::

::: {.footer}
Note that $y$ is standardized but the dummy code is left 0 or 1.
:::

## Effect sizes

- We can also use $R^2$ or Adjusted $R^2$ as before

```{r}
#| echo: true
#| collapse: true

r2(fit2, ci = 0.95)
```
::: {.fs70}
> "Around 18% of the variance in penguins' body mass was explained by sex,<br>R² = .181, 95% CI: [.110, .259], Adjusted R² = .178."

:::

## Estimating means by hand {.fs80}

$$
\begin{align}
\text{Model Equation: }&\hat{y}=b_0 + b_1(C_1)\\
&\hat{y} = 4545.68 -683.41(C_1)\\ \\
\text{Male penguins: }&\hat{y} = 4545.68 -683.41(0)=4545.68 \\
\text{Female penguins: }&\hat{y} = 4545.68 -683.41(1)=3862.27
\end{align}
$$

::: {.fragment}
- Each penguin $i$ will have a residual $e_i$ that pushes its body mass above or below the mean for its group (i.e., sex)

$$y_i=4545.68-683.41(C_1)+e_i$$
:::

## Estimating means in R

```{r}
#| echo: true
#| collapse: true

gmeans <- estimate_means(fit2, by = "sex")
gmeans
```

## Plotting means in R

```{r}
#| echo: true
#| eval: false

plot(gmeans)
```

```{r}
#| echo: false

plot(
  gmeans, 
  point = list(size = 3, color = "red"),
  pointrange = list(size = 1, linewidth = 1)
) + 
  mytheme(24)
```

# Polytomous <br>Dummy Coding

::: {.section-art}
{{< lif "../../icons/question.json" colors=secondary:#2a76dd class=rc >}}
:::

## Coding many categories

- With three categories, we need two code variables: $[C_1, C_2]$

- The reference group will get $0$ on **all** code variables

- Each other group will get $1$ on a single code variable

::: {.columns .fs70 .tc}
::: {.column width="33%"}
*Alcohol as Reference*
<table style="border-collapse: collapse; margin-left: auto; margin-right: auto;">
<tr class="bb">
<th>Condition</th>
<th>C<sub>1</sub></th>
<th>C<sub>2</sub></th>
</tr>
<tr>
<td>Alcohol*</td>
<td>0</td>
<td>0</td>
</tr>
<tr>
<td>Placebo</td>
<td>1</td>
<td>0</td>
</tr>
<tr>
<td>Control</td>
<td>0</td>
<td>1</td>
</tr>
</table>
:::
::: {.column width="33%"}
*Placebo as Reference*
<table style="border-collapse: collapse; margin-left: auto; margin-right: auto;">
<tr class="bb">
<th>Condition</th>
<th>C<sub>1</sub></th>
<th>C<sub>2</sub></th>
</tr>
<tr>
<td>Alcohol</td>
<td>1</td>
<td>0</td>
</tr>
<tr>
<td>Placebo*</td>
<td>0</td>
<td>0</td>
</tr>
<tr>
<td>Control</td>
<td>0</td>
<td>1</td>
</tr>
</table>
:::
::: {.column width="33%"}
*Control as Reference*
<table style="border-collapse: collapse; margin-left: auto; margin-right: auto;">
<tr class="bb">
<th>Condition</th>
<th>C<sub>1</sub></th>
<th>C<sub>2</sub></th>
</tr>
<tr>
<td>Alcohol</td>
<td>1</td>
<td>0</td>
</tr>
<tr>
<td>Placebo</td>
<td>0</td>
<td>1</td>
</tr>
<tr>
<td>Control*</td>
<td>0</td>
<td>0</td>
</tr>
</table>
:::
:::

## Selecting the reference group

The reference group has implications for interpretation

1. **The reference group should be a useful comparison**
    + e.g., a control group or standard treatment

2. **The reference group should be well-defined**
    + e.g., not an "other" or "wastebasket" category
    
3. **The reference group should not be relatively small**
    + e.g., not a group of 10 when others have 100

## Creating dummy codes by hand

- To create the dummy codes, we can use `recode()` again

```{r}
#| echo: true
#| collapse: true

penguins$species_c1 <- recode(penguins$species, Adelie=0, Chinstrap=1, Gentoo=0)

penguins$species_c2 <- recode(penguins$species, Adelie=0, Chinstrap=0, Gentoo=1)
```

::: {.fragment}
- This can get tedious as the number of groups increases

- It's easier and less error-prone to use the factor shortcut

```{r}
#| echo: true
#| collapse: true

penguins$species <- factor(
  penguins$species,
  levels = c("Adelie", "Chinstrap", "Gentoo")
)
```
:::

## Regression with many dummy codes

- To use many dummy codes, we need **multiple regression**
    + Basically, we add another predictor and another slope
    
$$\hat{y} = b_0 + b_1C_1 + b_2C_2$$

- The intercept $b_0$ is the mean of the reference group
- The slope $b_1$ is the reference vs. $C_1=1$ group difference
- The slope $b_2$ is the reference vs. $C_2=1$ group difference

::: {.footer}
We will elaborate on multiple regression during the next lecture
:::

## Regression with many dummy codes

```{r}
#| echo: true
#| collapse: true

fit4 <- lm(body_mass ~ species_c1 + species_c2, data = penguins)
fit4
```

- *Adelie* species averages 3700.66 g
- *Chinstrap* species averages 32.43 g more than *Adelie* species
- *Gentoo* species averages 1375.35 g more than *Adelie* species

## Parameter significance testing

```{r}
#| echo: true
#| collapse: true

model_parameters(fit4)
```

- Intercept: *Is the Adelie species body mass zero?*

- $C_1$ Slope: *Is the Chinstrap-Adelie group difference zero?*

- $C_2$ Slope: *Is the Gentoo-Adelie group difference zero?*



## The factor shortcut

```{r}
#| echo: true
#| collapse: true

fit5 <- lm(body_mass ~ species, data = penguins)
model_parameters(fit5)
```

- Note that we already set the factor up on slide 25

- The parameter table includes which level each code is for

## Model significance testing

- What is the overall effect of species (à la ANOVA)?
    + Do all categories have the same mean?
    + We can actually give the `lm()` results to `aov()`

```{r}
#| echo: true
#| collapse: true

fit5b <- aov(fit5)
model_parameters(fit5b)
```

::: {.footer}
The linear model (lm) subsumes ANOVA as well!
:::

## Effect sizes {.fs80}

```{r}
#| echo: true
#| collapse: true

model_parameters(fit5, standardize = "refit")

r2(fit5, ci = 0.95)
```

- *Chinstrap* species averages 0.04 SDs higher body mass than *Adelie*

- *Gentoo* species averages 1.72 SDs higher body mass than *Adelie*

- Species explained 67% of the variance in body mass

## Estimating means by hand

$$
\begin{align}
\text{Equation: }&\hat{y} = b_0 + b_1(C_1) + b_2(C_2) \\
  &\hat{y} = 3700.66 + 32.43(C_1) + 1375.35(C_2) \\ \\
\text{Adelie: }   &\hat{y} = 3700.66 + 32.43(0) + 1375.35(0) = 3700.66 \\
\text{Chinstrap: }&\hat{y} = 3700.66 + 32.43(1) + 1375.35(0) = 3733.09 \\
\text{Gentoo: }   &\hat{y} = 3700.66 + 32.43(0) + 1375.35(1) = 5076.01
\end{align}
$$

## Estimating means in R

```{r}
#| echo: true
#| collapse: true

gmeans <- estimate_means(fit5, by = "species")
gmeans
```

## Plotting means in R

```{r}
#| echo: true
#| eval: false

plot(gmeans)
```

```{r}
plot(
  gmeans, 
  point = list(size = 3, color = "red"), 
  pointrange = list(size = 1, linewidth = 1)
) + 
  mytheme(24)
```

## Contrasts

- What about the difference between Chinstrap and Gentoo?

- How can we compare non-reference groups to each other?

```{r}
#| echo: true
#| collapse: true

estimate_contrasts(fit5, contrast = "species")
```
