---
title: "Lecture 10b Activity"
subtitle: "Unit {{< meta unit >}} · Chapter {{< meta chapter >}}"
format:
  html:
    css: ../../styles.css
    toc: false
---

![](../../img/conversation_2780E3.svg){.page-art}

*An in-class activity. Nothing to turn in and no answer key -- this one needs
other people, which is why it happens in class rather than at home.*

**The idea.** Once you have seen which way deleting a case pushes your
$p$-value, you can no longer make an unbiased decision about deleting it. The
room writing rules *before* the reveal is the only way to demonstrate that
honestly.

**Format.** About 15 minutes · threes · one laptop per group.

## 1. Look, then stop (4 minutes)

```{r}
#| eval: false

library(tidyverse)
library(easystats)

yp <- read_csv("yearspubs.csv")
fit <- lm(n_cites ~ n_pubs + yrs_since, data = yp)

check_outliers(fit)         # the default verdict
plot(check_outliers(fit))   # every case's influence, seen at once
```

The default threshold reports **no outliers**. But look at the plot: one case
sits far beyond every other -- about five times the influence of the
runner-up.

**Do not run anything else yet.**

## 2. Write your rule (5 minutes)

On a sheet the group signs, write a rule you would be willing to preregister:

- What counts as an outlier here? Name the statistic and the threshold, and say
  whether you accept a rule that flags nothing in this dataset.
- Under what circumstances would you remove a flagged case? Under what
  circumstances keep it?
- What will you report either way?

Hand the sheet to the front. No changes after this point.

## 3. The reveal (6 minutes)

Ask the same function for the other rule from the lecture. Cook's distance asks
how much a case *moves the model*; Mahalanobis distance asks how unusual its
predictor values are. Same data, same function, both at their own defaults.

```{r}
#| eval: false

outliers <- check_outliers(fit, method = "mahalanobis")
outliers      # this rule does flag something
yp[24, ]      # who is it?

yp_drop <- yp[!as.numeric(outliers), ]   # remove it, exactly as in the slides
fit_drop <- lm(n_cites ~ n_pubs + yrs_since, data = yp_drop)
compare_parameters(fit, fit_drop)
```

- Apply your own rule. What does your group report?
- Read out two or three rules from the front. **Does the room's answer depend on
  whose rule you use?** That is the finding.
- Now the honest question, and answer it to each other: having seen the reveal,
  would you have written the same rule five minutes ago?
- The package default said there was nothing to see. What does that tell you
  about using a default as your decision rule?

*If you have more time: run `check_model()` on four models breaking four
different assumptions and name each from its plot before running any test.*

{{< dstart summary="If you were not in class" >}}

The influential case is a researcher with 69 publications and 13 years since
their PhD -- a real, plausible person and the most influential observation in the
dataset. Removing that single row moves `n_pubs` from $+0.20$ to $-0.04$ and
`yrs_since` from $p = .10$ to $p = .04$. **One row decides which predictor the
paper is about**, and reasonable rules written by different groups will disagree
about whether to keep it.

That disagreement is the point, and it is not a failure of the room. It is what
"researcher degrees of freedom" means, made visible: several defensible analysts
with the same data and different, honestly-held rules produce different papers.
The protection is not finding the one true threshold; it is fixing your rule
before the reveal and reporting the analysis both ways.

Note also what the two rules did. Cook's distance at its default threshold of
0.8 reported nothing; Mahalanobis distance at *its* default flagged this case
immediately. Neither is wrong -- they measure different things, influence
versus unusualness in the predictors -- but a room that had run only the first
one would have concluded there was nothing here to decide. A default is a
convention someone else chose, not a finding.

{{< dstop >}}
