---
lecture: "b"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.t-title}
::: {.t-eyebrow}
{{< meta course >}}
:::

# Confidence Intervals

::: {.t-rule}
:::

::: {.t-meta}
Unit {{< meta unit >}} · Chapter {{< meta chapter >}} · Lecture {{< meta chapter >}}{{< meta lecture >}}
:::

::: {.t-attrib}
{{< meta attribution >}}
:::

![]({{< meta unit-art >}}){.t-art}
:::

```{r setup}
#| echo: false
#| message: false

# Installs the replace_path source hook, which rewrites the data path shown on
# the slide without changing the path the render actually reads. Quarto does not
# pick _common.R up on its own; without this line every `replace_path` option in
# this deck is a silent no-op. The chunk is *named* so it does not shift the
# unnamed-chunk-N numbering that the committed figure filenames depend on.
source("../../_common.R")
```


```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(patchwork)
library(distributional)
library(ggdist)
mytheme <- theme_gray(base_size = 30) + theme(panel.grid.minor = element_blank())
# Make mytheme the default rather than relying on every plot remembering
# `+ mytheme`. Plots built by other packages -- plot(), check_model(),
# ggeffects -- never get a manual `+ mytheme` and were rendering at
# ggplot's 11pt default, which reads as tiny next to the 40px slide text.
theme_set(mytheme)
```


# Types of Estimates

::: {.section-art}
{{< lif "../../icons/question.json" colors=secondary:#2a76dd class=rc >}}
:::

## Point and Interval Estimates {.fs80}


In statistical inference, we will often want **two** types of estimates...

::: {.columns}
::: {.column}
#### [Point Estimates]{.blue}
- What is our ***single best guess*** for the <br>value of the population parameter?

- This is usually the sample statistic

- Says nothing about uncertainty

*e.g., I don't know the average height of all adult men, but I will guess it equals the average height in my sample*
:::
::: {.column}
#### [Interval Estimates]{.blue}
- What is a ***range of reasonable guesses*** for the value of the population parameter?

- Usually the statistic plus/minus an amount reflecting uncertainty

*e.g., Given sampling error, any number within 2.9 cm of my sample average is also a reasonable guess*
:::
:::

::: {.footer}

:::

## Reasonable values


- Our uncertainty comes from the fact that sampling error causes different samples to yield different estimates

- The sampling distribution shows what statistic values would be more or less common across many such samples

- Reasonable values would be those where the sampling distribution is densest (i.e., the values that are the most commonly observed across different samples)

- With less uncertainty (e.g., larger samples), the range of reasonable values will be tighter/narrower


::: {.footer}

:::

## Sampling distribution


- But we don't have access to the sampling distribution!
    + This is because we usually only take one sample...

- So we will need to *estimate* the statistic values for which <br>the sampling distribution is densest
    + There are many different techniques for doing this

- We will use the frequentist **parametric** approach
    + This approach is very common and assumes that <br>the sampling distribution is *normally distributed*
    + What does that mean? Is it reasonable to assume?

::: {.footer}

:::

# The Normal Distribution

::: {.section-art}
{{< lif "../../icons/normal_curve.json" state=in-reveal colors=secondary:#2a76dd class=rc >}}
:::

## The normal distribution {.fs80}


::: {.columns}
::: {.column}
Also called the *Bell Curve* or the *Gaussian* distribution

We write "$x$ is normal with mean $\mu$ and SD $\sigma$" as

$$x \sim \text{Normal}(\mu, \sigma)\\x \sim \mathcal{N}(\mu,\sigma)$$

- $\mu$ is the mean, $\sigma$ is the SD

$\mathcal{N}(0,1)$ is the *standard normal*
:::
::: {.column}
```{r}
#| fig-width: 6
#| fig-height: 6
data.frame(mu = 0, sigma = 1) %>%
  ggplot(aes(y = "", dist = dist_normal(mu, sigma))) + 
  stat_dist_slab(fill = NA, color = "black") +
  coord_cartesian(expand = FALSE) +
  scale_color_viridis_c() + 
  labs(
    y = NULL,
    x = "Normal(0, 1) distribution"
  ) +
  mytheme
```
:::
:::

::: {.footer}

:::

## The mu parameter


Changing $\mu$ shifts the distribution left ($-$) or right ($+$)

```{r}
#| fig-width: 12
#| fig-height: 5
data.frame(mu = seq(-3, 3, 1)) %>% 
  ggplot(aes(y = "", dist = dist_normal(mu, 1), color = mu)) + 
  stat_dist_slab(fill = NA, linewidth = 1) +
  coord_cartesian(expand = FALSE) +
  scale_color_viridis_c(breaks = c(-3, 0, 3)) + 
  labs(
    y = NULL,
    x = "Normal(mu, 1) distribution",
    color = "mu"
  ) +
  mytheme
```


::: {.footer}

:::

## The sigma parameter


Changing $\sigma$ expands ($+$) or contracts ($-$) the distribution

```{r}
#| fig-width: 12
#| fig-height: 5
data.frame(sigma = seq(1, 5, 1)) %>% 
  ggplot(aes(y = "", dist = dist_normal(0, sigma), color = sigma)) + 
  stat_dist_slab(fill = NA, linewidth = 1) +
  coord_cartesian(expand = FALSE) +
  scale_color_viridis_c(breaks = c(1, 3, 5)) + 
  labs(
    y = NULL,
    x = "Normal(0, sigma) distribution",
    color = "sigma"
  ) +
  mytheme
```

::: {.footer}

:::

## The empirical rule


```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 3.4

sd_labels <- c(
  "1" = "68.3% within 1 SD",
  "2" = "95.4% within 2 SD",
  "3" = "99.7% within 3 SD"
)
emp_curve <- expand_grid(k = 1:3, x = seq(-4, 4, length.out = 600)) |>
  mutate(y = dnorm(x))

ggplot(emp_curve, aes(x, y)) +
  geom_area(data = filter(emp_curve, abs(x) <= k), fill = "#357EDD", alpha = 0.45) +
  geom_line(linewidth = 1) +
  facet_wrap(~k, labeller = labeller(k = sd_labels)) +
  scale_x_continuous(breaks = seq(-4, 4, 2)) +
  scale_y_continuous(breaks = NULL, expand = expansion(c(0, 0.05))) +
  labs(x = NULL, y = NULL)
```

~68.3\% of the Normal distribution falls within 1 SD of the mean<br>
~95.4\% of the Normal distribution falls within 2 SD of the mean<br>
~99.7\% of the Normal distribution falls within 3 SD of the mean

::: {.footer}

:::

## Quantiles in R


How many SDs contain a given percentage?

- We want the *middle* 95%, so we cut the 5% in half

- We thus calculate the 2.5th and 97.5th percentiles

- `qnorm()` does this for the Normal distribution

```{r}
#| echo: true
#| collapse: true
qnorm(0.025)
qnorm(0.975)
```
- So 95% of the Normal distribution is within ~1.96 SDs


::: {.footer}

:::

## Putting it all together {.fs80}


1. Our interval should cover the densest part of the sampling distribution

1. 95% of the density of a normal distribution is within ~1.96 SD of the mean

1. The SD of the sampling distribution is the standard error (SE) of the statistic

1. So, if we assume the sampling distribution is normally distributed, then...

1. Our interval estimate is the sample statistic plus/minus 1.96 * SE

$$\text{CI}_{95} = \bar{x} \pm (1.96 \times s_{\bar{x}})$$

This is the [95% Confidence Interval (CI)]{.b .blue}, and more on it soon

But this parametric approach all hinges on the assumption of normality (#4)...

::: {.footer}

:::

# Central Limit Theorem

::: {.section-art}
{{< lif "../../icons/converge.json" state=in-reveal colors=secondary:#2a76dd class=rc >}}
:::

## Non-normally distributed data


What happens to the sampling distribution of $\bar{x}$ <br>when $x$ itself is not normally distributed?

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4.0

# Population is Beta(2, 1): a ramp on [0, 1], clearly non-normal, mean 2/3.
set.seed(2025)
clt_mean <- 2 / 3
clt_sd <- sqrt(2 / 36)

clt_plot <- function(sizes, reps = 20000) {
  draws <- map_dfr(sizes, \(n) tibble(
    n = n,
    xbar = replicate(reps, mean(rbeta(n, 2, 1)))
  ))
  normal <- map_dfr(sizes, \(n) tibble(
    n = n,
    x = seq(0, 1.2, length.out = 400),
    y = dnorm(x, clt_mean, clt_sd / sqrt(n))
  ))
  ggplot(draws, aes(xbar)) +
    geom_histogram(aes(y = after_stat(density)), bins = 40,
                   fill = "#357EDD", alpha = 0.55) +
    geom_line(data = normal, aes(x, y), linewidth = 1) +
    facet_wrap(~n, labeller = labeller(n = \(v) paste("Sample Size =", v))) +
    scale_y_continuous(breaks = NULL, expand = expansion(c(0, 0.05))) +
    labs(x = "Sample Mean", y = NULL)
}

clt_plot(c(1, 2))
```

::: {.footer}

:::

## Central limit theorem


The sampling distribution of $\bar{x}$ will become more normal <br>as $n$ increases (even if $x$ itself is not normally distributed)

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 4.0

clt_plot(c(4, 8))
```

::: {.footer}

:::

## Normal approximations


- The CLT **applies to means and sums**
    + And to statistics built from them
    + Why? A mean sums many independent observations
    + Some values will be higher and others will be lower
    + These deviations will tend to cancel each other out
    + Central values are common and extreme values are rarer

::: {.fragment}
- Thus, we often assume a normal sampling distribution...
    + This becomes more reasonable as $n$ increases
:::

::: {.footer}

:::


# Confidence Intervals

::: {.section-art}
{{< lif "../../icons/question.json" colors=secondary:#2a76dd class=rc >}}
:::

## Calculations


- To estimate the 95% Confidence Interval of the mean

$$\text{CI}_{95}= \bar{x}\pm\left(1.96\times\frac{\sigma}{\sqrt{n}}\right)$$

::: {.fragment}
- However, this requires that we know $\sigma$, the population SD

- Usually we estimate $\sigma$ using $s$, the sample SD

- This estimation introduces a bit more uncertainty...
:::

::: {.footer}

:::

## A complication {.fs80}


- To account for this extra uncertainty, we need to...
    + Stop using the Normal distribution and `qnorm()`
    + Instead use the Student $t$ distribution and `qt()`

::: {.fragment}
- The $t$ distribution has more uncertainty when $n$ is small
    + It becomes increasingly normal as $n$ increases
    + The $t$ distribution has the same $\mu$ and $\sigma$ parameters
    + But it also has "degrees of freedom" $(\textit{df}=n-1)$
    + As $n$ increases, *df* increases, and $t$ becomes more normal
    
:::

::: {.footer}

:::

## The *df* parameter


As *df* increases, $t(\textit{df},\mu,\sigma)$ becomes increasingly normal

```{r}
#| fig-width: 12
#| fig-height: 5
standist::visualize("student_t(1, 0, 1)", "student_t(3, 0, 1)",
                    "normal(0, 1)", xlim = c(-6, 6)) +
  scale_color_discrete(breaks = c("student_t(1, 0, 1)", "student_t(3, 0, 1)", "normal(0, 1)")) +
  labs(color = NULL, y = NULL, x = NULL) +
  mytheme + theme(legend.position = "bottom")
```

::: {.footer}
Sometimes you'll also see *df* referred to as $\nu$
:::

## Comparing multipliers


When $n$ is small, the $t$ multiplier is higher (more uncertainty)

As $n$ increases, the multipliers become increasingly similar

```{r}
#| fig-width: 12
#| fig-height: 5
tibble(
  n = seq(2, 30, 1),
  Normal = qnorm(0.975),
  t = qt(0.975, n - 1)
) |> 
  pivot_longer(
    cols = -n, 
    names_to = "distribution", 
    values_to = "multiplier"
  ) |> 
  ggplot(aes(x = n, y = multiplier, color = distribution, shape = distribution)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  scale_y_continuous(breaks = seq(2, 14, 2)) +
  labs(x = "Sample Size (n)", color = NULL, shape = NULL) +
  mytheme
```

::: {.footer}

:::

## A correction to our formula


::: {.columns}
::: {.column}
#### Population SD is known

$$\text{CI}_{\%}=\bar{x}\pm z \left(\frac{\sigma}{\sqrt{n}}\right)$$

```{r}
#| echo: true
#| collapse: true

# n = 20
(z <- qnorm(0.975))

# n = 100
(z <- qnorm(0.975))
```

:::
::: {.column}
#### Population SD is estimated
$$\text{CI}_{\%}=\bar{x}\pm t \left(\frac{s}{\sqrt{n}}\right)$$

```{r}
#| echo: true
#| collapse: true

# n = 20
(t <- qt(0.975, df = 20 - 1))

# n = 100
(t <- qt(0.975, df = 100 - 1))
```
:::
:::

::: {.footer}

:::

## An example in R


```{r}
#| echo: true
#| collapse: true
#| message: false
#| replace_path: ["../../data/", ""]

tratings <- read_csv("../../data/teaching_ratings.csv")
summary(tratings$eval)
```

::: {.columns}
::: {.column .fragment}
Estimate mean and SE
```{r}
#| echo: true
#| collapse: true

x <- tratings$eval
n <- length(x)
xbar <- mean(x)
s <- sd(x)
se <- s / sqrt(n)

xbar
se
```
:::
::: {.column .fragment}
Estimate confidence interval
```{r}
#| echo: true
#| collapse: true

t <- qt(0.975, df = n - 1)
ci_lo <- xbar - (t * se)
ci_hi <- xbar + (t * se)

ci_lo
ci_hi
```

:::
:::

::: {.footer}
I will teach you easier ways to calculate CIs in R soon.
:::

# Practical Matters

::: {.section-art}
{{< lif "../../icons/toolbox.json" state=in-reveal colors=secondary:#2a76dd class=rc >}}
:::

## Reporting a CI {.fs80}


::: {.columns}
::: {.column}
- Always report point and interval estimates together

- In text, use this format:
    + Statistic = Point Estimate, X% CI: [Lower Bound, Upper Bound]
    + $\bar{x}=3.91$, 95% CI $[3.81, 4.00]$

- You can also put point and interval estimates in a table or a figure (though be sure to specify that they are X% CIs)
:::
::: {.column .tc}
*Example Table*

```{r}
tibble(
  Parameter = "Mean Eval",
  Estimate = 3.91,
  `95% CI` = "[3.81, 4.00]"
) |>  knitr::kable()
```

*Example Figure*
```{r}
#| fig-width: 5
#| fig-height: 2
ggplot(
  tibble(param = "", est = 3.91, lo = 3.81, hi = 4.00), 
  aes(x = est, xmin = lo, xmax = hi, y = param)
) + 
  geom_point(size = 4) + 
  geom_errorbar(width = 0.25, size = 2) + 
  coord_cartesian(xlim = c(3.5, 4.5)) +
  labs(x = "Mean Eval (with 95% CI)", y = NULL) +
  theme_gray(base_size = 24) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(),
        axis.ticks.y = element_blank())
```
:::
:::

::: {.footer}

:::

## Comparing CIs {.fs80}


- A *higher* confidence level will result in a *wider* interval
   + To claim more confidence, the interval must "pay" with width
   
::: {.columns}
::: {.column}
```{r}
#| echo: true
#| collapse: true
library(lsr)

ciMean(x, conf = 0.99)

ciMean(x, conf = 0.95)

ciMean(x, conf = 0.90)
```

:::
::: {.column}
```{r}
#| fig-width: 5
#| fig-height: 4

ggplot(
  tibble(level = c("0.90", "0.95", "0.99"), 
         est = c(3.906, 3.906, 3.906), 
         lo = c(3.827, 3.812, 3.781), 
         hi = c(3.985, 4.000, 4.031)), 
  aes(x = est, xmin = lo, xmax = hi, y = level)
) + 
  geom_point(size = 4) + 
  geom_errorbar(width = 0.25, linewidth = 1.5) + 
  coord_cartesian(xlim = c(3.7, 4.1)) +
  labs(x = "Estimated Mean Eval", y = "Confidence Level") +
  theme_gray(base_size = 24) +
  theme(panel.grid.minor = element_blank())
```
:::
:::

::: {.footer}

:::

## Interpreting CIs


It is common to misinterpret what confidence intervals mean

:::{.callout-important title="Incorrect Interpretation"}
"There is a 95% probability that the population parameter value lies within the 95% CI."
:::

The above is incorrect because it does not align with the frequentist view of probability

::: {.footer}
Bayesian interval estimates (credible intervals) actually can be interpreted this way
:::

## Interpreting CIs


The correct frequentist interpretation is quite a mouthful

:::{.callout-tip title="Correct Interpretation"}
"If we repeated the experiment over and over again (with different samples of the same size) and computed the 95% CI in each sample, then 95% of those intervals would contain the population parameter value."
:::

Alternatives that are reasonable shortcuts

:::{.callout-note title="Acceptable Interpretations"}
"We can be 95% confident that the 95% CI contains the population mean."

"The 95% CI contains highly reasonable estimates of the population mean."
:::

::: {.footer}
The first alternative isn't my favorite as it implies a view of probability as belief (which isn't very frequentist)
:::

## Another example


Let's estimate the mean sleep satisfaction of all grad students

We have a sample of 16 grad students that we can use

```{r}
#| echo: true
#| collapse: true
#| message: false
#| replace_path: ["../../data/", ""]

sleep <- read_csv("../../data/sleep.csv")

mean(sleep$satisfaction)

ciMean(sleep$satisfaction, conf = 0.95)
```

*"We estimate that the mean sleep satisfaction among all grad students is 68.75, 95% CI: [57.78, 79.72]. Thus, our best guess for this value is 68.75 but, given sampling error, any number from 57.78 to 79.72 would also be a highly reasonable guess."*

::: {.footer}

:::
