---
lecture: "a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.t-title}
::: {.t-eyebrow}
{{< meta course >}}
:::

# Comparing Two Groups

::: {.t-rule}
:::

::: {.t-meta}
Unit {{< meta unit >}} · Chapter {{< meta chapter >}} · Lecture {{< meta chapter >}}{{< meta lecture >}}
:::

::: {.t-attrib}
{{< meta attribution >}}
:::

![]({{< meta unit-art >}}){.t-art}
:::

```{r setup}
#| echo: false
#| message: false

# Installs the replace_path source hook, which rewrites the data path shown on
# the slide without changing the path the render actually reads. Quarto does not
# pick _common.R up on its own; without this line every `replace_path` option in
# this deck is a silent no-op. The chunk is *named* so it does not shift the
# unnamed-chunk-N numbering that the committed figure filenames depend on.
source("../../_common.R")
```


```{r}
#| echo: false
#| message: false

library(tidyverse)
library(distributional)
library(ggdist)
library(lsr)
mytheme <- theme_gray(base_size = 30) + theme(panel.grid.minor = element_blank())
# Make mytheme the default rather than relying on every plot remembering
# `+ mytheme`. Plots built by other packages -- plot(), check_model(),
# ggeffects -- never get a manual `+ mytheme` and were rendering at
# ggplot's 11pt default, which reads as tiny next to the 40px slide text.
theme_set(mytheme)
```


# Introductions

::: {.section-art}
{{< lif "../../icons/two_people.json" state=in-reveal colors=secondary:#2a76dd class=rc >}}
:::

## Group Comparisons

- Many research questions are about **comparing groups**

- Do two (or more) groups have the same population mean?
    + These groups may be naturally occurring<br />
      (e.g., Do *adults* or *children* watch more TV?)
    + Or they may correspond to experimental conditions<br />
      (e.g., Do outcomes differ in *treatment* and *placebo*?)
    + Two groups of measurements may be related/paired<br />
      (e.g., Is a person's memory better *before* or *after* exercise?)

## Types of Groups

- We will collect a sample of data from each group

- [Independent Samples:]{.b .blue} Observations are unrelated
    + e.g., Each participant only occurs in a single group/sample
    + e.g., Different people are assigned to condition A or B
  
- [Dependent/Paired Samples:]{.b .blue} Observations are linked
    + e.g., Each participant occurs in multiple groups/samples
    + e.g., Each person participates in both condition A and B

## Types of Comparisons

::: {.columns}
::: {.column}
[Significance Testing]{.b .blue}

- Is the difference between the two group means zero?

- The result is a test statistic, <br>$p$-value, and decision to retain or reject the null

- *Adults watch significantly more TV than do children, $t(99)=20.14, p<.001$*
:::
::: {.column .fragment}
[Estimation]{.b .blue}

- What are reasonable values for the difference between the two group means?

- The result is a point estimate and interval estimate (CI)

- *Adults watch $3.2$ more hours of TV per day than do children, 95% CI: $[1.8, 4.6]$*
:::
:::

# NHST Approaches

::: {.section-art}
{{< lif "../../icons/two_routes.json" state=in-reveal colors=secondary:#2a76dd class=rc >}}
:::

## Recap of Steps in NHST

1. Set **null and alternative hypotheses**, Type I Error Rate $(\alpha)$

1. Choose an appropriate test and calculate its **test statistic**

1. Estimate **null sampling distribution** for test statistic

1. Calculate **critical values** from null sampling distribution

1. Compare the observed test statistic to these critical values

1. Calculate $\boldsymbol{p}$**-value** and decide to **retain or reject** the null

## Independent Samples $t$-test

- **Research Question:** Do tenure track (G1) or non-tenure track (G2) professors receive better teaching evals, on average?

- Each professor is G1 or G2, so these are *independent* groups

::: {.fragment}
- Are the G1 and G2 samples drawn from populations with <br>the same mean (null) or different means (alternative)?

$$
H_0: \mu_1=\mu_2\\
H_1: \mu_1\neq\mu_2
$$
:::

## Example data

```{r}
#| echo: true
#| collapse: true
#| message: false
#| replace_path: ["../../data/", ""]

library(tidyverse)
tratings <- read_csv("../../data/teaching_ratings.csv")
tratings
```

## Data Exploration

```{r}
#| fig-height: 5

ggplot(tratings, aes(x = factor(track), y = eval)) +
  ggbeeswarm::geom_quasirandom(
    color = "cornflowerblue", 
    size = 3,
    width = 0.25
  ) +
  scale_y_continuous(limits = c(1, 5)) +
  scale_x_discrete(
    breaks = 1:2,
    labels = c("1 (Tenure)", "2 (Non-tenure)")
  ) +
  labs(
    x = "Track",
    y = "Eval"
  ) +
  mytheme
```

## Test Statistic
- If the null is true and $\mu_1 = \mu_2$, then $\mu_1 - \mu_2 = 0$

- We can use this as the basis of our test statistic

- But again, the math is easier if we standardize it

$$t_{diff}=\frac{\bar{x}_1 - \bar{x}_2}{SE_{diff}}$$

- So now we just need to estimate the SE of the difference

## Two approaches

- There are two main approaches to estimating $SE_{diff}$

- [Student's approach]{.b .blue} assumes "equal variance"
    + It assumes that both populations have the same SD
    + In parameter terms, it assumes that $\sigma_1 = \sigma_2$
    + This approach simplifies the math, so we start here
    
- [Welch's approach]{.b .blue} does not make this assumption
    + It has a more complicated equation
    + But it is usually more realistic

## Student's approach

- We can estimate $\sigma$ by "pooling" $s_1$ and $s_2$

$$
s_p = \sqrt{\frac{s_1^2(n_1-1) + s_2^2(n_2-1)}{n_1+n_2-2}}
$$

::: {.fragment}
- We can now use $s_p$ to estimate $SE_{diff}$

$$SE_{diff} = s_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}$$
:::

::: {.footer}
Pooling is taking a weighted average (weighted by sample size minus 1).
:::

## Student's approach

- Our test statistic for Student's approach is:

$$t_{diff} = \frac{\bar{x}_1 - \bar{x}_2}{s_p\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}}$$

- The null sampling distribution is assumed to be:

$$t_{diff} \sim t(\textit{df}=n_1+n_2-2, \mu=0, \sigma=1)$$

## Welch's approach

$$SE_{diff} = \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}$$

$$
t_{diff} \sim t\left(
\textit{df}=\frac{\left(\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}\right)^2}
{\frac{\left(\frac{s_1^2}{n_1}\right)^2}{n_1-1} + \frac{\left(\frac{s_2^2}{n_2}\right)^2}{n_2-1}}, 
\mu=0, 
\sigma=1\right)
$$


## Both approaches in R

- We can have R do all this math for us using `t.test()`
    + `formula`: in the format *outcome ~ group*
    + `data`: tibble containing *outcome* and *group*
    + `var.equal`: TRUE (Student's) or FALSE (Welch's)

## Student's test results {.fs80}

```{r}
#| echo: true
#| collapse: true
#| message: false

fit_s <- t.test(
  formula = eval ~ track,
  data = tratings,
  var.equal = TRUE # <1>
)

library(easystats)
model_parameters(fit_s)
```
1. Be sure to set `var.equal=TRUE` to use Student's approach

::: {.fragment .pv2}
"A Student's independent samples *t*-test found that tenure-track professors' average eval (M=3.87) was significantly lower than non-tenure-track professors' average eval (M=4.12), *t*(92)=–1.99, *p*=.049. The difference was estimated to be –0.25, 95% CI: [–0.51, 0.00]."
:::

::: {.footer}
Note that this result is *barely* significant.
:::

## Welch's approach in R {.fs80}

```{r}
#| echo: true
#| collapse: true

fit_w <- t.test(
  formula = eval ~ track,
  data = tratings,
  var.equal = FALSE # <1>
)

model_parameters(fit_w)
```
1. Be sure to set `var.equal=FALSE` to use Welch's approach

::: {.fragment .pv2}
"A Welch's independent samples *t*-test found that tenure-track professors' average eval (M=3.87) was significantly lower than non-tenure-track professors' average eval (M=4.12), *t*(21.62)=–2.19, *p*=.040. The difference was estimated to be –0.25, 95% CI: [–0.50, –0.01]."
:::

## Paired groups

- Sometimes observations in our groups are related
    + Obs. X in G1 is related somehow to obs. X in G2
    + e.g., repeated measures of the same person or thing
    
- If we ignore these relationships in our modeling...
    + Our SE estimates will tend to be too high (when the pairing correlation is positive)
    + So our $p$-values and CIs are wrong, costing power

## Paired samples $t$-test {.fs80}

- To fix this, we can calculate the paired differences

$$\Delta_i = x_{1i} - x_{2i}$$

- Then we do a one-sample $t$-test on the differences

$$t_\Delta = \frac{\bar{\Delta}}{SE_{\bar{\Delta}}} = \frac{\bar{\Delta}}{s_{\Delta}/\sqrt{n}}$$

$$t_\Delta \sim t(\textit{df}=n-1, \mu=0, \sigma=1)$$

## Example data

```{r}
#| echo: true
#| collapse: true
#| message: false
#| replace_path: ["../../data/", ""]

social <- read_csv("../../data/socialproblems.csv")
social
```


## Paired $t$-test in R

- We can use `t.test()` again but our formula changes

```{r}
#| echo: true
#| collapse: true
fit_p <- t.test(
  formula = Pair(endpoint, baseline) ~ 1,
  data = social
)

model_parameters(fit_p)
```

::: {.footer}
Note that $\textit{df}=33$ (not $66$, as two independent samples would give) because we count each pair only once.
:::

## One-sample $t$-test equivalence

```{r}
#| echo: true
#| collapse: true

# Calculate paired differences
social$change <- social$endpoint - social$baseline

# One-sample t-test on paired differences
fit_o <- t.test(social$change, mu = 0)

model_parameters(fit_o)
```

Note that these results are identical

# Effect Sizes

::: {.section-art}
{{< lif "../../icons/caliper.json" state=in-reveal colors=secondary:#2a76dd class=rc >}}
:::

## Effect sizes

- The $p$ value combines the effect size and uncertainty

- But sometimes we want to separate these quantities out

- A *pure* measure of effect size would be nice sometimes

- In the current case of group comparisons...
    + **How large is the difference between group means?**
    + Is this difference large enough to matter in practice?
    + Can we make it easier to interpret and compare?

## Group difference effect sizes

- The difference in means $(\bar{x}_1 - \bar{x}_2)$ is a simple effect size
    + This is an "unstandardized" effect size (in original units)
    + If the original units are intuitive, this might work well
    
- It may help comparison to standardize this value
    + The difference is now in *standard deviation* units
    
$$ES = \frac{\bar{x}_1 - \bar{x}_2}{SD}$$

::: {.footer}
The question now becomes... which SD to use for standardization?
:::

## Independent samples

- When comparing the means of independent samples...
    + We can use the pooled SD in the denominator
    
$$d = \frac{\bar{x}_1 - \bar{x}_2}{s_p} = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{\frac{(n_1 - 1)s_1^2+(n_2-1)s_2^2}{n_1+n_2-2}}}$$

- This effect size measure is often referred to as Cohen's $d$

- Other options include Hedge's $g$ and Glass' $\Delta$

::: {.footer}
$g$ corrects for small sample sizes and $\Delta$ uses the SD from one group only
:::

## Unstandardized difference

```{r}
#| echo: true
#| collapse: true
#| warning: false
model_parameters(fit_s)
```

::: {.pv2}
Tenure track professors' average teaching evaluations were 0.25 units lower (on a scale from 1 to 5) than were non-tenure track professors', 95% CI: [–0.51, 0.00].
:::

## Cohen's $d$

```{r}
#| echo: true
#| collapse: true

cohens_d(eval ~ track, data = tratings)
```

::: {.pv2}
Tenure track professors' average teaching evaluations were 0.56 standard deviations lower than were non-tenure track professors', 95% CI: [–1.12, 0.00].
:::

## Dependent samples

- When comparing the means of dependent/paired samples...
    + The denominator uses the SD of the paired differences
    
$$ES = \frac{\bar{x}_1 - \bar{x}_2}{s_{\Delta}}$$

- This is often referred to as a paired Cohen's $d$

::: {.footer}
Because it standardizes by the SD of the *differences*, paired $d$ is not directly comparable to independent-samples $d$
:::

## Example

```{r}
#| echo: true
#| collapse: true

model_parameters(fit_p)

cohens_d(Pair(endpoint, baseline) ~ 1, data = social)
```

Patients' social problems were 0.42 units (0.66 change-score SDs) lower at the endpoint than at baseline.
