---
lecture: "b"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.t-title}
::: {.t-eyebrow}
{{< meta course >}}
:::

# Comparing Many Groups

::: {.t-rule}
:::

::: {.t-meta}
Unit {{< meta unit >}} · Chapter {{< meta chapter >}} · Lecture {{< meta chapter >}}{{< meta lecture >}}
:::

::: {.t-attrib}
{{< meta attribution >}}
:::

![]({{< meta unit-art >}}){.t-art}
:::

```{r setup}
#| echo: false
#| message: false

# Installs the replace_path source hook, which rewrites the data path shown on
# the slide without changing the path the render actually reads. Quarto does not
# pick _common.R up on its own; without this line every `replace_path` option in
# this deck is a silent no-op. The chunk is *named* so it does not shift the
# unnamed-chunk-N numbering that the committed figure filenames depend on.
source("../../_common.R")
```


```{r}
#| echo: false
#| message: false

library(tidyverse)
library(distributional)
library(ggdist)
library(lsr)
mytheme <- theme_gray(base_size = 30) + 
  theme(panel.grid.minor = element_blank())
# Make mytheme the default rather than relying on every plot remembering
# `+ mytheme`. Plots built by other packages -- plot(), check_model(),
# ggeffects -- never get a manual `+ mytheme` and were rendering at
# ggplot's 11pt default, which reads as tiny next to the 40px slide text.
theme_set(mytheme)
```


# Introduction

::: {.section-art}
{{< lif "../../icons/question.json" colors=secondary:#2a76dd class=rc >}}
:::

## Rationale

- The $t$-test compares two groups' population means

- But sometimes we want to compare **three or more** groups

- This is what [one-way analysis of variance]{.b .blue} (ANOVA) does
    + One-way focuses on *one* source of variance (i.e., groups)
    + The one-way ANOVA is for **independent samples**/groups

$$
H_0: \text{it is true that } \mu_1=\mu_2=\mu_3 \cdots \\
H_1: \text{it is NOT true that } \mu_1=\mu_2=\mu_3 \cdots
$$


## Example dataset {.fs80}

```{r}
#| echo: true
#| collapse: true
#| message: false
#| replace_path: ["../../data/", ""]

library(tidyverse)
penguins <- read_csv("../../data/penguins.csv")
penguins
```

- Do the three penguin **species** have the same mean body mass?

## Exploratory analysis

```{r}
#| fig-height: 5

ggplot(penguins, aes(x = species, y = body_mass)) +
  ggbeeswarm::geom_quasirandom(
    color = "cornflowerblue", 
    size = 3,
    width = 0.25,
    alpha = 1/2
  ) +
  scale_y_continuous() +
  mytheme
```

## Total variance

*Across all observations (in all groups), what is the average squared difference between each observation and the global mean (calculated across all observations in all groups)?*

$$
\text{Var}(y)=\frac{1}{n}\sum_{i=1}^n(y_i - \bar{y})^2
$$

$n$ is the total sample size across all groups

$\bar{y}$ is the global mean of $y$ across all groups

## Sums of squares

- ANOVA uses [sums of squares]{.b .blue} ($SS$) rather than variances
    + $SS$ are the same as variances but don't divide by $n$
    + They are the **sums** of squared differences (not the means)

::: {.columns}
::: {.column}
$$\color{blue}{\text{SS}_{tot}=\sum_{i=1}^n(y_{i}-\bar{y})^2}$$
:::
::: {.column}
$$\text{Var}(y)=\frac{1}{n}\sum_{i=1}^n(y_{i} - \bar{y})^2$$
:::
:::

::: {.footer}
Using $SS$ yields a nicer interpretation and enables the use of the $F$ distribution
:::

## Between and within groups

- The total variation ($SS_{tot}$) contains two components

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 3.2

bw_groups <- tibble(group = factor(1:3), mu = c(-2.4, 0, 2.4))
bw_curves <- expand_grid(bw_groups, x = seq(-6, 6, length.out = 500)) |>
  mutate(y = dnorm(x, mu, 1)) |>
  expand_grid(panel = c("between", "within"))

# Between: arrows span the gaps between group means.
# Within: arrows span one SD either side of each group mean.
bw_arrows <- bind_rows(
  tibble(panel = "between",
         x = c(-2.4, 0, -2.4), xend = c(0, 2.4, 2.4),
         y = c(0.34, 0.30, 0.26)),
  tibble(panel = "within",
         x = bw_groups$mu - 1, xend = bw_groups$mu + 1, y = 0.30)
)

bw_labels <- c(
  between = "Between-group variation\n(differences among group means)",
  within  = "Within-group variation\n(deviations from group means)"
)

ggplot(bw_curves, aes(x, y)) +
  geom_line(aes(group = group), linewidth = 0.7, colour = "grey35") +
  geom_segment(data = bw_arrows, aes(x = x, xend = xend, y = y, yend = y),
               inherit.aes = FALSE, linewidth = 0.8,
               arrow = arrow(ends = "both", type = "closed",
                             length = unit(0.18, "cm"))) +
  facet_wrap(~panel, labeller = labeller(panel = bw_labels)) +
  # "group 1/2/3" as tick labels overlap at this base size; the axis title
  # carries the word instead.
  scale_x_continuous(breaks = bw_groups$mu, labels = 1:3) +
  scale_y_continuous(breaks = NULL, expand = expansion(c(0, 0.08))) +
  labs(x = "Group", y = NULL)
```

::: {.columns}
::: {.column}
*How different are the three groups/species (on average)?*
:::
::: {.column}
*How different are penguins <br>in the same group/species?*
:::
:::

## Decomposing the variation

- We can decompose total variation into two components:

::: {.columns}
::: {.column}
**Between-group variation**

$$\text{SS}_b = \sum_{k=1}^gn_k(\bar{y}_k - \bar{y})^2$$

*How far is each group mean $\bar{y}_k$ from the global mean $\bar{y}$?*
:::
::: {.column}
**Within-group variation**

$$\text{SS}_w = \sum_{k=1}^g\sum_{i=1}^{n_k}(y_{ik} - \bar{y}_k)^2$$

*How far is each observation $y_{ik}$ from its group mean $\bar{y}_k$?*
:::
:::

::: {.footer}
Note that the decomposition ensures that $SS_{tot}=SS_b + SS_w$
:::

## Testing for mean differences

- If $H_0$ is true, the groups' population means are the same
    + Thus, between-group variation $(SS_b)$ should be small

::: {.fragment}
- But how small is small enough? We need to contextualize it
    + ANOVA does so by comparing $SS_b$ to $SS_w$
    + It focuses on the ratio of $SS_b$ to $SS_w$ (i.e., $SS_b/SS_w$)
    + But we need to build in the group and sample sizes
    + Doing so will allow us to calibrate our uncertainty
:::


## Incorporating uncertainty

::: {.tc}
::: {.columns}
::: {.column}
**Mean Squares Between**

$$\textit{MS}_b = \frac{\textit{SS}_b}{g - 1}$$

:::
::: {.column}
**Mean Squares Within**

$$\textit{MS}_w = \frac{\textit{SS}_w}{n - g}$$
:::
:::

[The One-way ANOVA Test Statistic]{.b}
$$
F_{obs} = \frac{\textit{MS}_b}{\textit{MS}_w}\\
F_{obs} \sim F(\textit{df}_1 = g - 1, \textit{df}_2 = n - g)
$$
:::

::: {.footer}
$g$ is the number of groups, $n$ is the total sample size
:::

## The F distribution {.fs80}

::: {.columns}

::: {.column}

- The sum of squared *standard* normals has a $\chi^2$ distribution

- The ratio of two independent $\chi^2$ variables, *each divided by its df*, is $F$-distributed

- This is why $F$ uses $MS = SS/\textit{df}$

$$x \sim F(\textit{df}_1, \textit{df}_2)$$

:::

::: {.column}

```{r}
#| fig-width: 6
#| fig-height: 6

data.frame(df1 = 2, df2 = 339) |>
  ggplot(aes(y = "", dist = dist_f(df1, df2))) +
  stat_dist_slab(fill = NA, color = "black") +
  coord_cartesian(expand = FALSE) +
  scale_color_viridis_c() +
  labs(
    y = NULL,
    x = "F(2, 339) distribution"
  ) +
  mytheme
```

:::
:::

## The df1 parameter

```{r}
#| fig-width: 12
#| fig-height: 5

data.frame(df1 = seq(2, 10, 2)) |>
  ggplot(aes(y = "", dist = dist_f(df1, 30), color = df1)) + 
  stat_dist_slab(fill = NA, linewidth = 1) +
  coord_cartesian(expand = FALSE) +
  scale_color_viridis_c(breaks = c(2, 6, 10)) + 
  labs(
    y = NULL,
    x = "F(df1, 30) distribution",
    color = "df1"
  ) +
  mytheme
```

- Higher $\textit{df}_1$ has thinner tails (i.e., lower critical values)

## The df2 parameter

```{r}
#| fig-width: 12
#| fig-height: 5

data.frame(df2 = c(10, 20, 30, 40, 50)) |>
  ggplot(aes(y = "", dist = dist_f(10, df2), color = df2)) + 
  stat_dist_slab(fill = NA, linewidth = 1) +
  coord_cartesian(expand = FALSE) +
  scale_color_viridis_c(breaks = c(10, 30, 50)) + 
  labs(
    y = NULL,
    x = "F(10, df2) distribution",
    color = "df2"
  ) +
  mytheme
```

- Higher $\textit{df}_2$ has thinner tails (i.e., lower critical values)

# One-way ANOVA in R

::: {.section-art}
{{< lif "../../icons/question.json" colors=secondary:#2a76dd class=rc >}}
:::

## The aov function

- The `aov()` function will do this math for us

```{r}
#| echo: true
#| collapse: true

fit <- aov(
  formula = body_mass ~ species,
  data = penguins
)
```

::: {.fragment .pv4}
```{r}
#| echo: true
#| collapse: true

summary(fit)
```

:::

::: {.fragment}
```{r}
#| echo: true
#| collapse: true

# Critical value (to compare to F_obs = 343.6)
qf(.95, df1 = 2, df2 = 339)
```

:::

## Prettier output

```{r}
#| echo: true
#| collapse: true
#| message: false

library(easystats)
model_parameters(fit)
```

*We used a one-way ANOVA to compare the mean body masses of Adelie, Chinstrap, and Gentoo penguins and rejected the null hypothesis that these three groups have the same mean, $F(2,339)=343.63$, $p<.001$.*


## Estimating group means

- We can also estimate the model-implied group means

```{r}
#| echo: true
#| collapse: true
#| message: false

gmeans <- estimate_means(fit, by = "species")
gmeans
```

::: {.footer}
Note that this function requires the {emmeans} package to be installed
:::

## Plotting group means

```{r}
#| echo: true
#| eval: false

plot(gmeans)
```

```{r}
#| echo: false
#| eval: true

visualisation_recipe(
  gmeans, 
  jitter = list(size = 2, color = "red"), 
  line = list(linewidth = 1),
  pointrange = list(size = 0.8, linewidth = 1)
) |> 
  plot() +
  theme_gray(base_size = 20)
```

## An effect size for One-way ANOVA

- We conclude that $H_0$ is wrong, but *how wrong* was it?

- **How much of the total variation was between groups?**

$$\eta^2 = \frac{SS_b}{SS_{tot}}$$

- If $\eta^2=0$, then none of the variation was between groups

- If $\eta^2=1$, then all of the variation was between groups

## Application and interpretation {.fs80}

```{r}
#| echo: true
#| collapse: true

ss_b <- 146864214
ss_w <- 72443483
ss_tot <- ss_b + ss_w

ss_b / ss_tot
```

::: {.pv4}
```{r}
#| echo: true
#| collapse: true

eta_squared(fit)
```
:::

- So 67% of the variation in body mass was between groups/species

# Post-hoc Tests

::: {.section-art}
{{< lif "../../icons/question.json" colors=secondary:#2a76dd class=rc >}}
:::

## Many ways to reject the null

- Recall that our null hypothesis is that $\mu_1=\mu_2=\mu_3$

::: {.fs60}
| Possibility | $\mu_1=\mu_2$ | $\mu_1=\mu_3$ | $\mu_2=\mu_3$ | Hypothesis  |
| ----------- | ------------- | ------------- | ------------- | ----------- |
| 1           | TRUE          | TRUE          | TRUE          | null        |
| 2           | TRUE          | -             | -             | alternative |
| 3           | -             | TRUE          | -             | alternative |
| 4           | -             | -             | TRUE          | alternative |
| 5           | -             | -             | -             | alternative |
:::

::: {.footer}
Equality is transitive, so no other combinations are possible: if $\mu_1=\mu_3$ and $\mu_2=\mu_3$, then $\mu_1=\mu_2$
:::

- Thus, if we reject the null, we don't know exactly why...

## Pairwise $t$-tests

- To figure it out, we can run a $t$-test for each pair of groups
    + These are called "post hoc" (Latin for "after that") tests

$$\mu_1=\mu_2 \qquad \mu_1=\mu_3 \qquad \mu_2=\mu_3$$

- There are two main challenges with this approach
    + It will be a hassle to run all these tests as $g$ increases
    + Running many tests will increase the Type I Error rate

## Addressing the challenges

- For the first challenge, we will use programming
    + We will use R functions to do all the work for us

- For the second challenge, we will use two approaches
    + **Gating:** only run post-hoc tests if ANOVA is significant
    + $p$-value adjustment: increase $p$ based on \# of tests run

- We want to keep the overall ("familywise") error rate at $\alpha$
    + We will learn two common adjustment approaches

::: {.footer}
Gating is a common convention (adjustment alone also controls the familywise rate); we will follow it in this course
:::

## Bonferroni correction

- Multiply each $p$ by the number of tests run ($m$)

$$p' = p \times m$$

::: {.fs70}
| raw $p$     | $m$ | Bonferroni $p'$ |
| ----------- | --- | --------------- |
| .001 \*\*\* | 5   | .005 \*\*       |
| .005 \*\*   | 5   | .025 \*         |
| .019 \*     | 5   | .095            |
| .022 \*     | 5   | .110            |
| .103        | 5   | .515            |
:::

::: {.footer}
Bonferonni is simple but very conservative
:::

## Holm correction

- Pretend you do the tests from smallest to largest $p$-value

- Take the larger of $(p'_j=j\times p_j)$ and $(p'_{j+1})$

:::{.fs70}
| raw $p$     | rank $j$ | $p\times j$ | Holm $p'$ |
| ----------- | -------- | ----------- | --------  |
| .001 \*\*\* | 5        | .005        | .005 \*\* |
| .005 \*\*   | 4        | .020        | .020 \*   |
| .019 \*     | 3        | .057        | .057      |
| .022 \*     | 2        | .044        | .057      |
| .103        | 1        | .103        | .103      |
:::

::: {.footer}
Holm is complicated but more powerful
:::

## Post-hoc tests in R

::: {.fs90}
```{r}
#| echo: true
#| collapse: true

estimate_contrasts(
  model = fit, 
  contrast = "species", 
  method = "pairwise",
  p_adjust = "bonferroni"
)
```
:::


## Adjustment in R

::: {.fs90}
```{r}
#| echo: true
#| collapse: true

estimate_contrasts(
  model = fit, 
  contrast = "species", 
  method = "pairwise",
  p_adjust = "holm" # change the method here
)
```
:::
