---
lecture: "c"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.t-title}
::: {.t-eyebrow}
{{< meta course >}}
:::

# ANOVA Extensions

::: {.t-rule}
:::

::: {.t-meta}
Unit {{< meta unit >}} · Chapter {{< meta chapter >}} · Lecture {{< meta chapter >}}{{< meta lecture >}}
:::

::: {.t-attrib}
{{< meta attribution >}}
:::

![]({{< meta unit-art >}}){.t-art}
:::

```{r setup}
#| echo: false
#| message: false

# Installs the replace_path source hook, which rewrites the data path shown on
# the slide without changing the path the render actually reads. Quarto does not
# pick _common.R up on its own; without this line every `replace_path` option in
# this deck is a silent no-op.
source("../../_common.R")

library(tidyverse)
library(distributional)
library(ggdist)
library(lsr)
mytheme <- theme_gray(base_size = 30) + 
  theme(panel.grid.minor = element_blank())
# Make mytheme the default rather than relying on every plot remembering
# `+ mytheme`. Plots built by other packages -- plot(), check_model(),
# ggeffects -- never get a manual `+ mytheme` and were rendering at
# ggplot's 11pt default, which reads as tiny next to the 40px slide text.
theme_set(mytheme)
```


# Unequal Variances

::: {.section-art}
{{< lif "../../icons/uneven.json" colors=secondary:#2a76dd class=rc >}}
:::

## Welch's one-way test

- The ANOVA from last lecture used Student's approach
    + It simplifies the math by assuming all $\sigma$ are equal
    + The "homogeneity of variance" assumption
    
- But we can extend Welch's approach to ANOVA as well
    + This will let us relax this assumption
    + The math is more complex than we have time for
    + But I will show you how to implement it
    
## Example dataset

```{r}
#| echo: true
#| collapse: true
#| message: false
#| replace_path: ["../../data/", ""]

library(tidyverse)
penguins <- read_csv("../../data/penguins.csv")
penguins
```

## Exploratory analysis

```{r}
#| fig-height: 5

ggplot(penguins, aes(x = species, y = body_mass)) +
  ggbeeswarm::geom_quasirandom(
    color = "cornflowerblue", 
    size = 3,
    width = 0.25
  ) +
  scale_y_continuous() +
  mytheme
```

## Welch's test in R

```{r}
#| echo: true
#| collapse: true
#| message: false

library(easystats)

fit <- oneway.test(
  formula = body_mass ~ species,
  data = penguins
)

model_parameters(fit)
```

# Repeated Measures

::: {.section-art}
{{< lif "../../icons/repeat_loop.json" colors=secondary:#2a76dd class=rc >}}
:::

## One-way RM ANOVA

- To compare *independent* groups, we can use...
    + Independent samples $t$-test (for exactly 2 groups)
    + One-way ANOVA (for 3+ groups)

- To compare dependent/paired groups, we can use...
    + Paired samples $t$-test
    + [One-way Repeated Measures ANOVA]{.b .blue}

## Further partitioning

- We will partition the variation into *three* final components

```{mermaid}
%%| fig-width: 12

%%{ init: {
    "theme": "base",
    "themeVariables": { "fontSize": "24px" }
} }%%

flowchart LR
  T[Total Variation] --> B(Between-Group Variation)
  T --> W(Within-Group Variation)
  W --> S(Subject Variation)
  W --> E(Error Variation)
```

::: {.columns .pv2}
::: {.column .tc}
**Independent Samples**
$$F_{obs} = \frac{MS_b}{MS_w}$$

:::
::: {.column .tc}
**Dependent Samples**
$$F_{obs} = \frac{MS_b}{MS_e}$$
:::
:::

## Partitioning the variation

$$\text{SS}_b = \sum_{k=1}^g n_k(\bar{y}_k - \bar{y})^2 \qquad \text{SS}_w = \sum_{k=1}^g\sum_{i=1}^{n_k}(y_{ik} - \bar{y}_k)^2$$

::: {.columns .fs60}
::: {.column}
*How far are group means from the grand mean?*
:::
::: {.column}
*How far are scores from their group means?*
:::
:::


$$\quad \color{red}{\text{SS}_s = g \sum_{i=1}^{n} (\bar{y}_i - \bar{y})^2 \qquad \quad \text{SS}_e = \text{SS}_w - \text{SS}_s} \qquad \qquad$$

::: {.columns .fs60}
::: {.column}
*How far are subject means from the grand mean?*
:::
::: {.column}
*How much variation is not yet explained?*
:::
:::

## The One-way RM ANOVA

::: {.tc .b}
Mean Squares and Degrees of Freedom
:::

$$\text{MS}_b = \frac{\text{SS}_b}{g - 1} \qquad \text{MS}_e = \frac{\text{SS}_e}{(n-1)(g-1)}$$

::: {.tc .b}
Test Statistic and Null Sampling Distribution
:::

$$
F_{obs} = \frac{\text{MS}_b}{\text{MS}_e} \\
F_{obs} \sim F(\textit{df}_1=g-1, \textit{df}_2 = (n-1)(g-1))
$$

## Example data

```{r}
#| echo: true
#| collapse: true
#| message: false
#| replace_path: ["../../data/", ""]

depression <- read_csv("../../data/depression.csv")
depression
```

## Exploratory analysis

```{r}
#| fig-height: 5

ggplot(depression, aes(x = time, y = hrsd)) +
  ggbeeswarm::geom_quasirandom(
    color = "cornflowerblue", 
    size = 3,
    width = 0.25
  ) +
  scale_y_continuous() +
  mytheme
```

## Fitting the model in R

::: {.fs90}
```{r}
#| echo: true
#| collapse: true
#| message: false

library(afex)
fit <- aov_ez(
  id = "patient",
  dv = "hrsd",
  data = depression,
  within = "time"
)
model_parameters(fit)
```
:::

::: aside
*Note.* `id` names the column identifying who each row came from. A factor goes in `within` when everyone is measured at every level, and in `between` when each person appears at only one.
:::

## Estimating means

```{r}
#| echo: true
#| collapse: true
#| warning: false
#| message: false

gmeans <- estimate_means(fit, by = "time")
gmeans
```

## Plotting means

```{r}
#| echo: true
#| eval: false

plot(gmeans)
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

visualisation_recipe(
  gmeans, 
  jitter = list(size = 2, color = "red"), 
  line = list(linewidth = 1),
  pointrange = list(size = 0.8, linewidth = 1)
) |> 
  plot() +
  theme_gray(base_size = 24)
```


## The partial eta-squared effect size

- How much of the *non-subject* variation is between groups?

$$\eta_{partial}^2 = \frac{\text{SS}_b}{\text{SS}_b + \text{SS}_e}$$

::: {.fragment}
```{r}
#| echo: true
#| collapse: true

eta_squared(fit)
```
:::

## Post-hoc tests {.fs80}

- We can average pairwise time-contrasts over patients
- e.g., what is the T1--T0 difference *for the average patient?*

```{r}
#| echo: true
#| collapse: true
#| warning: false
#| message: false

estimate_contrasts(
  model = fit,
  contrast = "time",
  method = "pairwise"
)
```

# Translation Activity

::: {.section-art}
{{< lif "../../icons/translate.json" colors=secondary:#2a76dd class=rc >}}
:::

## Example 1 (Social Psych)

::: {.fs70}
> I read about a theory that says that a common enemy can unite political rivals. To test this, I have 200 American participants first complete a measure of hostility toward members of the opposing political party. They then read a news article that describes a foreign government threatening the USA. After answering some basic questions about the article, the participants complete another measure of hostility toward members of the opposing political party. Both measures of hostility are continuous and range from $-5$ to $5$.
:::

1. What might the data frame look like for this study?

2. What might my null and alternative hypotheses be?

3. What statistical test(s) from this course might I use here?


## Example 2 (Clinical Psych)

::: {.fs70}
> I read about a theory that says that the personality trait of "Openness to Experience" (from the Big 5) is related to psychosis such that people with higher levels of this personality trait experience more symptoms of psychosis (e.g., hallucinations, delusions, and disorganized thinking) on average. To explore this idea, I collect data from 150 members of my local community, including self-report measures of the Big 5 personality traits and psychotic symptoms. All these measures are continuous and range from $0$ to $100$.
:::

1. What might the data frame look like for this study?

2. What might my null and alternative hypotheses be?

3. What statistical test(s) from this course might I use here?

## Example 3 (Sports Science)

::: {.fs70}
> I read about a new pre-workout stretching technique that is supposed to dramatically reduce cramping. To see if it really works, I instruct one-third of the players on the soccer team I coach to use the new technique at the start of practice, another third to use a traditional stretching technique, and the final third to not stretch at all. After practice, I collect reports from each player on how much cramping they experienced during practice (on a scale from $1$ to $10$).
:::

1. What might the data frame look like for this study?

2. What might my null and alternative hypotheses be?

3. What statistical test(s) from this course might I use here?

## Small-Group Activity

**Step 1.** Write down a few of your own examples (e.g., 2 to 4). The hypotheses should be testable using one-sample $t$-tests, two-sample $t$-tests, one-way tests, or correlations.

**Step 2.** Form small groups of two to three students and take turns reading each others' examples. Discuss the answers to our three questions and brainstorm together.

1. What might the data frame look like for this study?

2. What might my null and alternative hypotheses be?

3. What statistical test(s) from this course might I use here?
