---
title: "Lecture 02b Assignment"
format:
  html:
    css: ../../styles.css
    toc: false
---

![](../../img/exams_2780E3.svg){width=300, height=300}

## Instructions

*Create a new Quarto document (HTML format), clear out the boilerplate code/text, and use this document to store your answers to each of the three questions below. Be sure to enable `embed-resources` using the header below. If your answer involves writing text, then use markdown to write 2--4 sentences per question; if you answer involves running code, then use one or more code chunks. For questions that ask you to fix code, correct any errors (while achieving the stated goal) and include only the working code in your document.*

```markdown
---
title: "Lecture 02b Assignment"
name: "Your Name"
format: html
embed-resources: true
---
```

*Please clearly mark which part of your document is answering which question (e.g., using markdown headings) and remove any "scratch work" or failed attempts from your document before submitting it. Finally, note that many essay-style questions will require you to read between the lines and use your imagination (as opposed to just looking up the answer in the lecture).*

## Question 1 (Assignment and Naming)

a) Why is it important to give your objects in R good names? When might poorly thought-out names cause problems for a programmer?

b) Sam is designing a treehouse. The plan is basically a big rectangular box with a wall that is 5.32 ft wide and 4.74 ft long (Sam is very precise). Using R, calculate the area of the treehouse and save that value to an object named `area`. Then use that object to calculate the `volume` of the treehouse, assuming the walls are 5.61 ft tall. Print the `volume` object.

$$Area = Length \times Width$$

$$Volume = Area \times Height$$

c) Help me fix my code! "I want to assign a name to my extraversion score on the Five Factor Model personality test."

```{r, renv.ignore=TRUE}
#| error: true
#| collapse: true

5FM_extraversion <- 23
```

d) Help me fix my code! "I want to calculate the average of three objects corresponding to negative emotion self-reports."

```{r}
#| error: true
#| collapse: true

anger <- 3
sadness <- 4
fear <- 3
negative_emotion <- (anger + sadness + faer) / 3
```

## Question 2 (Functions)

a) What are functions and arguments (in the context of programming) and why are they useful?

b) Help me fix my code! "I want to round 1/16 to three decimal places."

```{r, renv.ignore=TRUE}
#| error: true
#| collapse: true

round(1 / 16; places = 3)
```

## Question 3 (Vectors)

a) What did I mean when I said that "functions typically either transform each element in a vector or summarize all elements in a vector"? In your answer, provide an example (in text) of when each of these behaviors might be useful.

b) Below is a table of the top 5 countries' production (i.e., [GDP Per Capita at PPP](https://www.worldometers.info/gdp/gdp-by-country/)) in 2017. Save the numbers only to a vector and then use R to create a new vector that contains the log-transformed versions of each number. Print that new vector.

{{< dstart summary="Hint" >}}
Your output should look like this:

```{r}
#| echo: false

gdp <- c(19.485, 12.238, 4.872, 3.693, 2.651)
log(gdp)
```

{{< dstop >}}

|Country|Trillions |
|-------|----------|
|USA    |19.485    |
|China  |12.238    |
|Japan  |4.872     |
|Germany|3.693     |
|India  |2.651     |

---

✅ Done! *Please save and then render your document to an HTML file. Check it over to make sure it is complete (and rendered properly) and then submit it as directed by your instructor. Note that, if any of your code chunks has an error in it, the HTML file will not render properly (this is why removing scratch work is important).*
