---
title: "Lecture 02b Practice"
subtitle: "Unit {{< meta unit >}} · Chapter {{< meta chapter >}}"
format:
  html:
    css: ../../styles.css
    toc: false
---

![](../../img/studying_2780E3.svg){.page-art}

*These activities are for practice only; there is nothing to turn in. Answer
keys are included, so try each question yourself before opening them.*

## Question 1

Which of the following commands do you think will create an error in R? Why?

a) `score@T1 <- 3.2`
b) `score_at_T1 <- 3.2`
c) `score at T1 <- 3.2`
d) `1_score <- 3.2`
e) `ScoreAtTime1 <- 3.2`

Make your best guesses just by looking at the commands and then check your guesses by running the commands in R. *Bonus:* Can you come up with a better name for this variable?

{{< dstart summary="Answer Key" >}}

| Code | Result (reason) |
|------|-----------------|
|`score@T1 <- 3.2`|🛑 Error (`@` not allowed in name)|
|`score_at_T1 <- 3.2`|✅ No Error|
|`score at T1 <- 3.2`|🛑 Error (spaces not allowed in name)|
|`1_score <- 3.2`|🛑 Error (name can't start with a number)|
|`ScoreAtTime1 <- 3.2`|✅ No Error|

{{< dstop >}}

## Question 2

a) Use the `sqrt()` function to calculate the square root of 3 and assign it to a variable called `y`. Print the value of `y` to the console.
b) Use a function to round `y` to the nearest whole number.
c) Use a function to round `y` to two decimal places.

{{< dstart summary="Answer Key" >}}

```{r}
#| collapse: true

# Part (a)
y <- sqrt(3)
y

# Part (b)
round(y)

# Part (c)
round(y, digits = 2)
```

{{< dstop >}}

## Question 3

What do you expect the result to be for the following commands?

a) `sqrt(4 + 5)`
b) `sqrt(4) + 5`

Make your best guesses just by looking at the commands and then check your guesses by running the commands in R.

{{< dstart summary="Answer Key" >}}

```{r}
#| collapse: true

# Part (a): Add then square root, same as sqrt(9)
sqrt(4 + 5)

# Part (b): Square root then add, same as 2 + 5
sqrt(4) + 5 
```

{{< dstop >}}

## Question 4

Imagine that you ran a stand selling bananas for three days. On the first day, you sold \$30 worth of bananas and spent \$15 on supplies. On the second day, you sold \$50 worth of bananas and spent \$15 on supplies. On the third day, you sold $40 worth of bananas and spent \$100 on repairs to the stand (due to an unfortunate fire).

a) Create two vectors named `sales` and `costs` to store how much you sold and how much you spent on each day, respectively.

b) Subtract the `costs` object from the `sales` object and save the result to a new object named `profits`. Print this to see your profits per day.

c) Use the `sum()` function to calculate your total profits over all three days. Was your effort *fruitful*, i.e., did you make money overall?

{{< dstart summary="Answer Key" >}}

```{r}
#| collapse: true

# Part (a)
sales <- c(30, 50, 40)
costs <- c(15, 15, 100)

# Part (b)
profits <- sales - costs
profits

# Part (c): No, I lost money!
sum(profits)
```

{{< dstop >}}
