---
lecture: "a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.t-title}
::: {.t-eyebrow}
{{< meta course >}}
:::

# Power Analysis

::: {.t-rule}
:::

::: {.t-meta}
Unit {{< meta unit >}} · Chapter {{< meta chapter >}} · Lecture {{< meta chapter >}}{{< meta lecture >}}
:::

::: {.t-attrib}
{{< meta attribution >}}
:::

![]({{< meta unit-art >}}){.t-art}
:::

```{r setup}
#| echo: false
#| message: false

library(tidyverse)
library(easystats)
library(patchwork)
mytheme <- function(sz = 30) {
  theme_gray(base_size = sz) + 
  theme(panel.grid.minor = element_blank())
}
# Make mytheme the default rather than relying on every plot remembering
# `+ mytheme`. Plots built by other packages -- plot(), check_model(),
# ggeffects -- never get a manual `+ mytheme` and were rendering at
# ggplot's 11pt default, which reads as tiny next to the 40px slide text.
theme_set(mytheme())
```

# Power Analysis

## Power analysis {.fs90}

::: {.columns}
::: {.column}
- **Power** is the probability of correctly rejecting the null hypothesis (given that it is false)
    + Power is closely related to<br>the effect size, $n$, and $\alpha$
    + These quantities are so related that we can estimate any one from the other three
    + **Power analysis** is doing these kind of calculations
    
:::
::: {.column .tc}
```{r}
#| echo: false
#| fig-width: 5.6
#| fig-height: 5.6

# Panel A is the reference; B-D each change one thing relative to it.
pwr_cfg <- tibble(
  panel = c("A", "B", "C", "D"),
  note  = c("reference", "smaller α", "larger ES", "larger N"),
  d     = c(2.6, 2.6, 3.8, 2.6),
  alpha = c(0.05, 0.005, 0.05, 0.05),
  se    = c(1, 1, 1, 0.65)
) |>
  mutate(cv = qnorm(1 - alpha), panel = factor(panel, levels = c("A", "B", "C", "D")))

pwr_grid <- expand_grid(pwr_cfg, x = seq(-4, 8, length.out = 800)) |>
  mutate(h0 = dnorm(x, 0, 1), h1 = dnorm(x, d, se))

# This figure sits in a narrow column, so it needs a smaller base size than the
# deck default: on-screen text is base_size/72 * displayed_px / fig_width_in.
ggplot(pwr_grid, aes(x)) +
  # power: the part of H1 beyond the critical value
  geom_area(data = filter(pwr_grid, x >= cv), aes(y = h1),
            fill = "#7FBFEA", alpha = 0.9) +
  # type I error: the part of H0 beyond the same point, drawn last and solid so
  # it reads as its own colour instead of blending to purple through the blue
  geom_area(data = filter(pwr_grid, x >= cv), aes(y = h0), fill = "#F92F60") +
  geom_line(aes(y = h0), linewidth = 0.7) +
  geom_line(aes(y = h1), linewidth = 0.7) +
  geom_vline(aes(xintercept = cv), linewidth = 0.4) +
  geom_text(data = pwr_cfg, aes(x = -3.9, y = 0.47, label = note),
            hjust = 0, vjust = 1, fontface = "italic", size = 5.5) +
  facet_wrap(~panel, ncol = 1, strip.position = "right") +
  scale_y_continuous(breaks = NULL, expand = expansion(c(0, 0.12))) +
  labs(x = NULL, y = NULL) +
  theme_gray(base_size = 17) +
  theme(panel.grid.minor = element_blank(),
        axis.text = element_blank(), axis.ticks = element_blank())
```

::: {.fs50 .gray}
Blue = power, red = Type I error rate (α), line = decision criterion
:::

:::
:::

## Types of power analysis {.fs90}

- [A priori]{.b .red}: determine $n$ given others
    + Useful for planning before a new data collection (primary use)
- [Sensitivity]{.b .blue}: determine minimum $\text{ES}$ given others
    + Useful when you are unsure of the effect size to expect
- [Post-hoc]{.b .gray}: determine power $(1-\beta)$ given others
    + Not very useful in practice (since your $\text{ES}$ is based on the data)
- [Criterion]{.b .gray}: determine $\alpha$ given others
    + Not very useful in practice (since $\alpha$ is usually fixed anyway)

## Power analysis techniques

- **Analytical** power analysis
    + Solve using equations (easier but less flexible)
    + e.g., the G*Power program, the {WebPower} package

::: {.pv2}
- **Simulation**-based power analysis
    + Solve using simulated data (harder but more flexible)
    + e.g., the Mplus program, the {simr} package

:::

## *t*-test power {.fs90}

- A $t$-test compares the means of two groups (independent or paired)
- To do a priori power analysis, we need an expected effect size
- The effect size that WebPower wants you to provide is Cohen's $d$

$$d=\frac{\bar{x}_2 - \bar{x}_1}{s}$$

- **Best Approach:** estimate $d$ from previous studies
- **Last Resort:** plan to detect a "small" effect $(d=0.2)$

## Example 1 {.fs90}

What sample size would provide 80% power to detect an effect of size *d*=0.3 when comparing two independent groups?

```{r}
#| echo: true
#| collapse: true
#| message: false

library(WebPower)
wp.t(
  d = 0.30,
  power = 0.80,
  alpha = 0.05,
  type = "two.sample"
)
```

We would thus need 175 participants in each group (for 350 total)


## Example 2 {.fs90}

Now, what sample size would provide 80% power to detect an effect of size *d*=0.5 when comparing two independent groups?

```{r}
#| echo: true
#| collapse: true
#| message: false

library(WebPower)
wp.t(
  d = 0.50,
  power = 0.80,
  alpha = 0.05,
  type = "two.sample"
)
```

We need *fewer* participants to detect a larger (more obvious) effect


## Example 3 {.fs90}

Now, what sample size would provide 80% power to detect an effect of size *d*=0.5 when comparing two *dependent* groups?

```{r}
#| echo: true
#| collapse: true
#| message: false

library(WebPower)
wp.t(
  d = 0.50,
  power = 0.80,
  alpha = 0.05,
  type = "paired"
)
```

We need *fewer* participants when each provides repeated measures

## One-way ANOVA power {.fs90}

- One-way ANOVA compares the means of multiple groups
- WebPower wants the $f$ effect size and $k$ (the number of groups)

$$f=\frac{\sigma_b}{\sigma_w}$$

- **Best Approach:** estimate $f$ from previous studies
- **Last Resort:** plan to detect a "small" effect $(f=0.10)$

::: {.footer}
Check out `wp.rmanova()` for repeated measured ANOVA power analysis
:::

## Example 1

What sample size would provide 80% power to detect an effect of size *f*=0.25 when comparing 4 independent groups?

```{r}
#| echo: true
#| collapse: true
#| message: false

wp.anova(
  f = 0.25,
  power = 0.80,
  alpha = 0.05,
  k = 4
)
```

We need 178 participants total (45 per group)

## Example 2 {.fs90}

What sample size would provide 80% power to detect an effect of size *f*=0.25 when comparing 5 independent groups?

```{r}
#| echo: true
#| collapse: true
#| message: false

wp.anova(
  f = 0.25,
  power = 0.80,
  alpha = 0.05,
  k = 5
)
```

With more groups, we need fewer participants per group (39)


## Linear model power {.fs90}

- We technically have two types of power for LM
    + Both will compare the $R^2$ values of "full" and "reduced" models
    
- **Omnibus power** (of the overall model's $F$ test)
    + Full model: Includes all predictors
    + Reduced model: Excludes all predictors
    
- **Targeted power** (of specific slopes' $t$ tests)
    + Full model: Includes all predictors
    + Reduced model: Excludes the targeted predictor(s)

## Linear model power {.fs90}

- We will use the $R^2$ values to calculate the $f^2$ effect size

$$f^2=\frac{R_{\text{full}}^2 - R_{\text{reduced}}^2}{1-R_{\text{full}}^2}$$

- **Last Resort:** plan to detect a "small" effect $(f^2=.02)$

- WebPower also needs the number of predictors in each model
    + $p_1$ is the number of predictors in the full model
    + $p_2$ is the number of predictors in the reduced model

## Omnibus example 1 {.fs90}

What sample size would provide 80% power to explain 10% of the variance in the outcome variable using 5 predictor variables?

- First, let's calculate $f^2$ from the specified information
    + Full model has all predictors, so $p_1=5$ and $R^2_{\text{full}}=0.10$
    + Reduced model has no predictors, so $p_2=0$ and $R^2_{\text{red}}=0.0$
    
```{r}
#| echo: true
#| collapse: true

r2_full <- 0.10
r2_red <- 0.00
f2 <- (r2_full - r2_red) / (1 - r2_full)
f2
```

## Omnibus example 1 {.fs90}

Now we can calculate the sample size needed in our example

```{r}
#| echo: true
#| collapse: true
#| message: false

wp.regression(
  p1 = 5,
  p2 = 0,
  f2 = 0.111,
  alpha = 0.05,
  power = 0.80
)
```

We would need around 121 total participants

## Omnibus example 2 {.fs90}

Now, what sample size would provide 80% power to explain 10% of the variance in the outcome variable using 7 predictor variables?

- First, let's calculate $f^2$ from the specified information
    + Full model has all predictors, so $p_1=7$ and $R^2_{\text{full}}=0.10$
    + Reduced model has no predictors, so $p_2=0$ and $R^2_{\text{red}}=0.0$
    
```{r}
#| echo: true
#| collapse: true

r2_full <- 0.10
r2_red <- 0.00
f2 <- (r2_full - r2_red) / (1 - r2_full)
f2
```

## Omnibus example 2 {.fs90}

Now we can calculate the sample size needed in our example

```{r}
#| echo: true
#| collapse: true
#| message: false

wp.regression(
  p1 = 7,
  p2 = 0,
  f2 = 0.111,
  alpha = 0.05,
  power = 0.80
)
```

We need a larger sample with more predictors (but the same $f^2$)

## Targeted example 1 {.fs90}

If sex and age explain 20% of the variance in our outcome variable, what sample size would provide 80% power to detect the effect of a third predictor that explains an additional 5%?

- Full model has three predictors, so $p_1=3$ and $R^2_{\text{full}}=0.25$
- Reduced model has two predictors, so $p_2=2$ and $R^2_{\text{red}}=0.20$
    
```{r}
#| echo: true
#| collapse: true

r2_full <- 0.25
r2_red <- 0.20
f2 <- (r2_full - r2_red) / (1 - r2_full)
f2
```

## Targeted example 1 {.fs90}

Now we can calculate the sample size needed in our example

```{r}
#| echo: true
#| collapse: true
#| message: false

wp.regression(
  p1 = 3,
  p2 = 2,
  f2 = 0.067,
  alpha = 0.05,
  power = 0.80
)
```

We need around 119 total participants


## Targeted example 2 {.fs90}

In the same example as the last one, what would change if sex and age had explained 40% of the variance instead of only 20%?

- Full model has three predictors, so $p_1=3$ and $R^2_{\text{full}}=0.45$
- Reduced model has two predictors, so $p_2=2$ and $R^2_{\text{red}}=0.40$
    
```{r}
#| echo: true
#| collapse: true

r2_full <- 0.45
r2_red <- 0.40
f2 <- (r2_full - r2_red) / (1 - r2_full)
f2
```

## Targeted example 2 {.fs90}

Now we can calculate the sample size needed in our example

```{r}
#| echo: true
#| collapse: true
#| message: false

wp.regression(
  p1 = 3,
  p2 = 2,
  f2 = 0.091,
  alpha = 0.05,
  power = 0.80
)
```

With more total variance explained, we need fewer participants
