---
lecture: "a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.t-title}
::: {.t-eyebrow}
{{< meta course >}}
:::

# Data Types & Packages

::: {.t-rule}
:::

::: {.t-meta}
Unit {{< meta unit >}} · Chapter {{< meta chapter >}} · Lecture {{< meta chapter >}}{{< meta lecture >}}
:::

::: {.t-attrib}
{{< meta attribution >}}
:::

![]({{< meta unit-art >}}){.t-art}
:::

## Roadmap: More Programming

::: {.columns .pv4}

::: {.column width="60%"}
1. Strings
  
2. Factors

3. Packages
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Strings

## Strings {.fs70}

::: {.columns .pv4}
::: {.column width="60%"}
-   When talking to R, we need a way to distinguish
    -   Object/function names (e.g., the *mean* function)
    -   Text/character data (e.g., the word *mean*)
    
::: {.fragment .mt1}
-   [Strings]{.b .blue} are R's way of storing text data
    -   Strings can store any characters (no rules!)
    -   Strings are created and displayed with [quotes]{.b .green}
:::
::: {.fragment .mt1}
-   R has great tools for working with strings
    -   Strings can be collected into vectors
    -   Special functions can transform strings
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/quotation.json" trigger=hover colors=secondary:#2a76dd class=rc >}}

::: {.fragment}
`name <- "John Doe"`
:::

:::
:::

## Strings Need Quotation Marks

```{r}
#| echo: true
#| error: true

# Without quotes, R looks for an object named red
my_color <- red
```

```{r}
#| echo: true

# With quotes, it is a string
my_color <- "red"
my_color
```

::: {.takeaway}
Quotation marks are what separate a **value** from a **name**. Without them R
goes looking for an object, which is why the error mentions one you never made.
:::

## Strings Can Hold Any Symbol

```{r}
#| echo: true

# Characters that are illegal in object names are fine inside a string
dye <- "red#40"
dye

dyes <- c("red#40", "blue#02")
dyes
```

::: {.takeaway}
Inside quotes the naming rules no longer apply --- spaces, `#`, and anything else
are just characters. R only reads what is inside as text.
:::

## Pitfall: Math Does Not Work on Strings

```{r}
#| echo: true
#| error: true

dyes + 1
```

```{r}
#| echo: true
#| warning: true

mean(dyes)
```

::: {.takeaway}
`mean()` does not stop with an error --- it warns and hands back `NA`. A result
that **looks like an answer** is easier to miss than a failure.
:::

## Functions That Do Work on Strings

```{r}
#| echo: true

# How many elements are in the vector?
length(dyes)

# How many characters are in each element?
nchar(dyes)

# Some functions only make sense for strings
toupper(dyes)
```

::: {.takeaway}
`length()` counts elements whatever they are; `nchar()` and `toupper()` only mean
anything for text. What a function accepts follows from what it is for.
:::

# Factors

## Factors {.fs70}

::: {.columns .pv4}
::: {.column width="60%"}
-   [Factors]{.b .blue} are used to represent categorical data
    -   Factors have multiple possible [levels]{.b .blue}
    -   Levels are [discrete]{.b .green} and mutually-exclusive

::: {.fragment .mt1}
-   Sometimes levels are numeric "codes"
    -   e.g., 1=Drama, 2=Action, 3=Comedy
    -   R needs to know these aren't numbers!
    -   We can even give each level a descriptive [label]{.b .green}
    -   Or we can just store the levels as strings
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/theater.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

## From Numeric Codes to a Factor

```{r}
#| echo: true

# Ask 10 kids to order 1: nuggets, 2: pizza, or 3: salad
food <- c(2, 2, 1, 2, 1, 2, 1, 1, 2, 2)

food2 <- factor(food)
food2

table(food2)
```

::: {.takeaway}
`factor()` tells R these numbers are labels, not quantities. The `Levels:` line is
how a factor prints --- and R found only the values that **actually occur**.
:::

## Declaring Levels That Did Not Occur

```{r}
#| echo: true

# Nobody ordered salad, but 3 was always a possible answer
food3 <- factor(food, levels = c(1, 2, 3))
food3

table(food3)
```

::: {.takeaway}
State the levels yourself and the count for salad is a meaningful `0` rather than
a category that silently does not exist.
:::

## Labelling the Levels

```{r}
#| echo: true

food4 <- factor(food, levels = c(1, 2, 3),
                labels = c("nuggets", "pizza", "salad"))
food4

table(food4)
```

::: {.takeaway}
`levels` are the values in your data; `labels` are what you want printed. Naming
them here means never again looking up what `2` meant.
:::

## Pitfall: Levels are Not Labels

```{r}
#| echo: true

# levels must match what is IN the data; labels are what you want printed
food5 <- factor(food, levels = c("nuggets", "pizza", "salad"),
                labels = c(1, 2, 3))
food5
```

::: {.takeaway}
All `<NA>`: R searched the data for the string `"nuggets"` and found only the
number `2`. It does not error --- it **quietly returns missing values**.
:::

## Levels Can Be Strings to Begin With

```{r}
#| echo: true

genre <- c("pop", "metal", "pop", "rock", "rap", "rap", "pop", "rock")

genre2 <- factor(genre)
genre2

table(genre2)
```

::: {.takeaway}
With no `levels` given, R sorts the observed values alphabetically --- which is
rarely the order you want in a table or a plot.
:::

# Packages

## Packages {.fs70}

::: {.columns .pv4}
::: {.column width="60%"}
-   [Cookbooks]{.b .green} are a great way to learn to cook
    -   *They contain lots of recipes and instructions*
    -   Browse an online **bookstore** for a cookbook
    -   **Order** it to add it to your personal **bookshelf**
    -   To use, **pull** the cookbook off the shelf

::: {.fragment .mt1}
-   [Packages]{.b .blue} are like cookbooks for R
    -   *They contain helpful functions and datasets*
    -   Browse an online **repository** for a package
    -   **Install** it to add it to your personal **library**
    -   To use, **load** the package from the library
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/books.json" trigger=hover colors=secondary:#2a76dd class=rc >}}

::: {.fragment}
`library("pkg_name")`
:::
:::
:::

## Pitfall: A Function R Cannot Find

```{r}
#| echo: true

students <- c("mary anne", "BENjamin", "Lee")
students
```

```{r}
#| echo: true
#| error: true

str_to_title(students)
```

::: {.takeaway}
The stringr package has a function that fixes capitalization --- but "could not
find function" means R cannot find that **name**. Check the spelling first, then
whether the function's package is installed and loaded.
:::

## Installing a Package {.fs70}

-   [RStudio > Extras pane > Packages tab > Install]{.b .blue} button

::: {.fragment .mt1}
-   Or run it yourself:
:::

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: false

install.packages("stringr")
```

::: {.fragment}
::: {.callout-warning}
Run this in the [Console]{.b}, never inside a Quarto document --- it would reinstall the package every time you render.
:::
:::

## Installing is Not Loading

```{r}
#| echo: true
#| error: true

str_to_title(students)
```

```{r}
#| echo: true
#| message: false

library("stringr")

str_to_title(students)
```

::: {.takeaway}
Installing puts the package on your [library]{.b .green} shelf; it is still not in
this session until you take it down with `library()`. Install once per computer
(and again after upgrading R); load **once per session**.
:::

## Keeping Packages Updated {.fs70}

-   [RStudio > Extras pane > Packages tab > Update]{.b .blue} button

::: {.fragment .mt1}
-   Worth doing occasionally, but [not]{.b .blue} in the middle of an analysis --- a package update can change results
:::

## Learning More: Vignettes

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: false

browseVignettes("stringr")
```

::: {.takeaway}
Many packages ship long-form articles showing how they are meant to be used ---
usually a better starting point than the help page for any one function. This
opens a browser window, so run it in the Console too.
:::
