---
title: "Lecture 01b Practice"
subtitle: "Unit {{< meta unit >}} · Chapter {{< meta chapter >}}"
format:
  html:
    css: ../../styles.css
    toc: false
---

![](../../img/studying_2780E3.svg){.page-art}

*These activities are for practice only; there is nothing to turn in. Answer
keys are included, so try each question yourself before opening them.*

## Question 1 (R and RStudio)

There are many ways to customize the RStudio interface. To explore the possibilities, go to **Tools** > **Global Options**. The most noticeable things you can change are aspects of RStudio's appearance. If you click on the *Appearance* section, you can select a different zoom level, editor font, font size, and editor theme. I usually teach with the default settings and you can keep everything the same if you want your computer to match the one I teach from, but for my own work I use the "Tomorrow Night 80s" theme and a custom font called [Fira Code](https://affcom.ku.edu/posts/firacode/). I also make slight changes to the *Pane Layout* section that I'd be happy to describe to you in class if you're interested.

All of this stuff is optional, but a few changes that I'd recommend making are to go to the *General* section and make sure that "Restore .RData into workspace at startup" is unchecked and that "Save workspace to .RData on exit" is set to "Never". Finally, under the *Code* section and the Display tab, I like to check the box for "Allow scroll past end of document".

## Question 2 (Console)

Read what the user wanted to do and look at the command they sent to R. Fix the errors in the commands and calculate the answers.

a)  I want to add 20 and 0.2 and divide that sum by 4.1. But I get an error when I try:<br />`[20 + .2] \ 4.1`
b)  I want to multiply 6 by the sum of 2 and 1,300. But I get an error when I try:<br />`6 x (2 + 1,300)`

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

Change the brackets `[]` to parentheses `()` because parentheses `()` are the order-of-operations operator in R. Also, change the backslash `\` to a forward slash `/` because forward slash `/` is the division operator in R.

```{r}
#| comment: "#>"
#| collapse: true

(20 + .2) / 4.1
```

**Answer (b)**

Change the `x` to an asterisk `*` because asterisk `*` is the multiplication operator in R. Also, remove the comma `,` from `1,300` because R doesn't use commas to delimit digits in large numbers.

```{r}
#| comment: "#>"
#| collapse: true

6 * (2 + 1300)
```

</blockquote></details></p>

## Question 3 (Scripts)

Create a new R script file and copy your answers to Question 2 into it. Add some comments to explain which line of code corresponds to 2a and 2b. It is a good idea to get into the practice of keeping your code organized and commented. Organized code will also make grading easier, and a happy grader is associated with a happy student. 😉

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>
```{r}
#| eval: false

#### File > New File > R Script (or equivalent shortcuts)

# Question 2a
(20 + .2) / 4.1

# Question 2b
6 * (2 + 1300)

#### File > Save (or equivalent shortcuts)
```
</blockquote></details></p>
